---
title: "Summarize pathology and dementia severity"
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
library(ggplot2)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r constants}
cohorts <- c("OHSU", "rush", "emory")


```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }
```



```{r define-input}
indir <- "2_Pipeline/01-Explore_data/"
files_patho <- sapply(cohorts, function(x) { paste0(indir, "patho_", x, ".txt") })
files_main <- sapply(cohorts, function(x) { paste0(indir, "phenos_cleaned_", x, ".txt") })


```


```{r define-output}

```


```{r read-data}
main <- sapply(files_main, fread)
patho <- sapply(files_patho, fread)

```


```{r prepare-data-1}
## Harmonize ID columns
setnames(patho$OHSU, "bfonum", "ID")
setnames(patho$rush, "projid", "ID")
main$emory[, ID := NULL]
setnames(main$emory, "GUID", "ID")
setnames(patho$emory, "GUID", "ID")


### Merge main phenos and new patho/severity phenos
merged <- sapply(cohorts, function(acohort) {
    merge(main[[acohort]], patho[[acohort]], by = "ID", all.x = TRUE)
})

## For OHSU, prepare the path_dx variable to a more readable form
merged$OHSU[, path_dx_merged := gsub("1.0\\)", "", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("1\\)", "", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("2.0\\)", "+", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("2.0", "+", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("3.0\\)", "+", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("2\\)", "+", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("3\\)", "+", path_dx_merged)]
merged$OHSU[, path_dx_merged := gsub("  ", " ", path_dx_merged)]


## Get sample sizes
sampsizes <- sapply(merged, nrow)

```

## Sample sizes

```{r }
Mkable(rbind(sample_size = sampsizes))
```


## Brain pathology variables

* Emory - none available
* OHSU - path_dx available in format like "1.0) AD" or "1.0) LBD 2.0) AD 3.0) VD"
* Rush - 
    + Braak, Reagan AD, NIA AD, chronic gross infarct, and chronic
      microinfarct scores available with full completeness
    + hspath_typ and cvda_4gp available but poor completeness

### Plots:
#### OHSU
For the plot label, I converted the codes from format like "1.0) AD"
or "1.0) LBD 2.0) AD 3.0) VD"
to "AD" or "LBD + AD + VD" to improve readability and to clean up
redundant labels. However, I am not
sure whether that is the correct way to interpret the codes, so we
should check this. 


```{r fig.width = 8}
par(mar = c(8, 4, 3, 2))
PlotCategorical("path_dx_merged", merged$OHSU, xsrt = 45)

```


#### Rush
```{r }
par(mar = c(5, 4, 3, 2))
PlotCategorical("braaksc", merged$rush)
PlotCategorical("ad_reagan", merged$rush)
PlotCategorical("niareagansc", merged$rush)
PlotCategorical("ci_num2_gct", merged$rush)
PlotCategorical("ci_num2_mct", merged$rush)
PlotCategorical("hspath_typ", merged$rush)
PlotCategorical("cvda_4gp2", merged$rush)
```

## Dementia severity measures

### Plots

#### OHSU

```{r }
par(mar = c(8, 4, 3, 2))
PlotCategorical("clinical_dx", merged$OHSU, xsrt = 45)
```


### Rush 
```{r }
par(mar = c(5, 4, 3, 2))
PlotCategorical("cogdx", merged$rush)

```

#### Emory

```{r }
PlotCategorical("UDS_dx", merged$emory)

```

```{r }

```




```{r finish}

```
