---
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: html_document
---


```{r goal, include = FALSE}
###### GOAL ######
```


```{r online-resources, include = FALSE}
## WGCNA package documentation
## https://www.rdocumentation.org/packages/WGCNA/versions/1.73

## one tutorial (there are many online)
## https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0

## hub gene selection
## https://support.bioconductor.org/p/69858/#70116
```


```{r load-packages, include = FALSE}
library(data.table)
library(argparse)
library(WGCNA)
library(ggplot2)
library(stringr)

rm(list = ls())
```


```{r functions, include = FALSE}
###### FUNCTIONS ######
source("~/comsv/Svattathil_Library/svattathil_functions.r")

PrepareDataME <- function(modulename, aphenodata, amedata) {
    ## Extract eigengene for current module and add protsample as column
    mes.to.merge <- data.frame(protsample=rownames(amedata), 
                               mod.eigen=amedata[,modulename])

    ## Add eigengene column to phenos
    merged <- merge(mes.to.merge, aphenodata, by = "protsample")

    ## Convert to data frame and add protsample as rownames
    samps <- merged$protsample
    setDF(merged)
    rownames(merged) <- samps
    return(merged)
}

GetCoefficientAndSE <- function(regression.out, type, var="miRNA") {
    if(type=="ordinal") {
        return(summary(regression.out)$coefficients[var, c("Value", "Std. Error")]) }
    else if(type %in% c("gaussian", "binomial", "lmm")) {
            return(summary(regression.out)$coefficients[var, c("Estimate", "Std. Error")])
        }
}

GetPvalue <- function(regression.out, type, var="miRNA") {
    if(type=="ordinal") {
        return(pnorm(abs(summary(regression.out)$coefficient[var, "t value"]),
                     lower.tail=FALSE) * 2) }
    else if(type %in% c("binomial", "neg.binomial")) {
                         summary(regression.out)$coefficients[var, "Pr(>|z|)"] }
    else if(type %in% c("gaussian", "lmm")) {
                             summary(regression.out)$coefficients[var, "Pr(>|t|)"] }
}

```
	
		
```{r setup, include = FALSE}
###### SETUP ######
### Command line arguments
## Create parser
parser <- ArgumentParser()
parser$add_argument("--cohort", type="character",
                    help="OHSU, rush, or emory", default = "emory")
parser$add_argument("--softpower", type="integer",
                    help="value for softpower", default = 9)
parser$add_argument("--cutheight", type="double",
                    help="value for mergeCutHeight", default = 0.30)
parser$add_argument("--network_set", type="character",
                    help="cohort-specific or consensus", default = "cohort-specific")

## Read from parser
args <- parser$parse_args()


### Constants
options(stringsAsFactors = FALSE)
source("1_Code/project_constants.r")
nps <- as.character(npsvars.bin)

```

```{r manual-definitions, include = FALSE}
### Choose a cutting height for defining sample outliers
## based on the sample dendrogram
if(args$cohort %in% c("OHSU", "emory")) {
    clust.height <- 0
}

if(args$cohort == "rush") {
    clust.height <- 180
}

```

---
title: "WGCNA for `r args$cohort` -- mergeCutHeight `r args$cutheight`" 
---


```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```


```{r word-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```


```{r html-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```


```{r input}
### Files that exist
datdir <- "2_Pipeline/02-Prepare_analysis_data/"
infiles <- list(
    resid        = paste0(datdir, "resid_regress_covars_", args$cohort, ".txt"),
    phenos       = paste0(datdir, "phenos_cleaned_SVs_", args$cohort, ".txt"),
    shared_genes = "2_Pipeline/04-Build_networks/shared_genes_for_consensus_module.txt"
)
```
	
	
```{r output}
### Files to be created
outdir <- paste0("2_Pipeline/04-Build_networks/Wgcna_out/",
                 ifelse(args$network_set == "cohort-specific",
                        "Cohort-specific/",
                        "Using_shared_genes/"))
MyMkdir(outdir)
MyMkdir(paste0(outdir, "/Module_genes/"))

runid <- paste0(args$cohort, "_softpower", args$softpower,
                "_cutheight", sprintf("%.2f", args$cutheight))
if(args$network_set == "consensus") {
    runid <- paste0("for_consensus_", runid)
}

outfiles <- lapply(list(
    conn_stats       = "_gene_conn_stats.txt",
    regress_stats    = "_eigengene_regress_stats.txt",
    hubgenes         = "_hub_genes.txt",
    n_genes          = "_module_gene_counts.txt", 
    eigengenes       = "_eigengenes.txt",
    sigmodules       = "_associated_modules.txt",
    background_genes = "_background_genes.txt", 
    log              = ".log",
    adj_df           = "_protein_adjacency_matrix.txt",
    adj_log          = "_protein_adjacency_matrix.log",
    tombase          = "_tom",
    protdat          = "_expr_data.txt"
), function(x) { paste0(outdir, runid, x) })

## Will also write full list of module genes for each module
Outfile_module_genes <-  function(module) {
    paste0(outdir, "Module_genes/", runid, "_module_genes_", module, ".txt") }
```


```{r read-data}
###### MAIN ######
### Read in data
resid.all <- fread(infiles[["resid"]])
phenos.all <- fread(infiles[["phenos"]])
if(args$network_set == "consensus") { 
    shared_genes <- scan(infiles[["shared_genes"]], what = "")
}
```


```{r prepare-data}
### Prepare data
## Restrict phenos to samples in resid data (because they did different filtering)
phenos <- phenos.all[protsample %in% colnames(resid.all), ]

## Restrict resid to samples in phenos and convert to
## matrix with one row per protein and one row per sample
resid <- as.matrix(resid.all[, phenos$protsample, with = FALSE])
rownames(resid) <- resid.all$protein

## Select proteins with data for at least 50 participants
tokeep <- apply(resid, 1, function(x) {sum(!is.na(x)) > 50 })
forzobs <- resid[tokeep, ]

## If preparing for consensus module,  filter to proteins shared across cohorts
if(args$network_set == "consensus") {
    forzobs <- forzobs[rownames(forzobs) %in% shared_genes, ]
}

## Z-scale
zobs <- t(apply(forzobs, 1, scale))
rownames(zobs) <- rownames(forzobs)
colnames(zobs) <- colnames(forzobs)

## Make sure proteins are sorted alphabetically
zobs <- zobs[sort(rownames(zobs)), ]

```


# WGCNA parameters

These I experimented with:

* softpower:         `r args$softpower`
* mergeCutHeight:    `r args$cutheight`


These I left the same across runs:

* deepSplit:         4
* networkType:       signed
* TOMType:           signed
* corType:           bicor
* maxPOutliers:      0.10
* reassignThreshold: 0
* pamRespectsDendro: FALSE

# Quality Control
The input data includes observations for `r pN(ncol(zobs))` participants and 
`r pN(nrow(zobs))` proteins. 

## Missingness
WGCNA has built-in filters for gene and sample missingness. 

```{r check-missingness, include=TRUE}
## Transpose to get matrix in shape that WGCNA expects,
## which is row per sample and column per protein
zobs.tr <- t(zobs)

##  Check data for excessive missing values and identification of outlier samples
gsg <- goodSamplesGenes(zobs.tr, verbose = 3);
gsg$allOK

## If any, remove the offending genes and samples from the data
if (!gsg$allOK) {
    if (sum(!gsg$goodGenes) > 0)
        printFlush(paste("Removing genes:", paste(colnames(zobs.tr)[!gsg$goodGenes],
                                                  collapse = ", ")))
    if (sum(!gsg$goodSamples) > 0)
        printFlush(paste("Removing samples:", paste(rownames(zobs.tr)[!gsg$goodSamples],
                                                    collapse = ", ")))

    zobs.tr.filtered <- zobs.tr[gsg$goodSamples, gsg$goodGenes]
}
if(gsg$allOK) {
    zobs.tr.filtered <- zobs.tr[gsg$goodSamples, gsg$goodGenes]
}
```

The missingness filters left `r pN(nrow(zobs.tr.filtered))` participants and 
`r pN(ncol(zobs.tr.filtered))` genes. 


```{r transpose}
## zobs.tr.filtered has one row per sample and column per protein
dotransp <- FALSE

if(!dotransp) {
    ## don't transpose to keep one row per sample and one column per protein
    dat0 <- zobs.tr.filtered
}

if(dotransp) {
    ## transpose to keep one row per protein and one column per sample
    dat0 <- t(zobs.tr.filtered)
}
```


## Visual check - outlier samples 
WGCNA includes a function to perform hierarchical
clustering on the expression data to visually identify outlier
samples. 

```{r check-sample-outliers, include="TRUE", results="hide"}
## Cluster elements to check for obvious outliers
## and plot the tree
obstree <- hclust(dist(dat0), method = "average");

par(cex = 0.6, mar = c(0,4,2,0))
plot(obstree, main = "Clustering to detect outliers", sub="",
     xlab="", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)

##** See clust.height definitions in MANUAL DEFINITIONS section **##

## Draw line for cut threshold
## and define sample clusters using specified cutHeight
abline(h = clust.height, col = "red")
clust <- cutreeStatic(obstree, cutHeight = clust.height, minSize = 10)
clusttab <- (table(clust))
```


```{r remove-outliers}
## Specify which cluster to keep
## filter data
## and then re-cluster samples
cluster.to.keep <- names(clusttab)[which.max(clusttab)]
keep_samples <- clust==cluster.to.keep
protdat <- dat0[keep_samples, ]
noutliers <- sum(!keep_samples)
obstree2 <- hclust(dist(protdat), method = "average")
```


Based on the tree shape, I chose a cut height of `r clust.height`. 
This removed `r noutliers` outlier samples, leaving 
`r nrow(protdat)` samples for analysis. 

## Visual check - phenotype correlation 
We can use the sample dendrogram to visually check for 
variables that correlate with sample groupings,
indicating potential batch or other confounding effects. 

```{r visualize-phenos-on-clusters, include=TRUE, eval = TRUE}
## Extract categorical variables
## convert each to a factor
phenos.basic <- phenos[protsample %in% rownames(protdat),
                       intersect(c("protsample", "Batch", "msex", "clinical_dx"),
                                 names(phenos)), with=FALSE]
sampnames <- phenos.basic$protsample
phenos.basic[, protsample := NULL] # remove protsample column after extracting sampnames
setDF(phenos.basic)
rownames(phenos.basic) <- sampnames

## Convert values to factors
for(apheno in names(phenos.basic)) { phenos.basic[,apheno] <- factor(phenos.basic[,apheno]) }

## Verify that rownames match between protein and trait data
all(rownames(phenos.basic) == rownames(protdat))

## Convert traits to a color representation:
## white means low, red means high, grey means missing entry
trait_colors <- sapply(colnames(phenos.basic), function(apheno) {
    labels2colors(phenos.basic[,apheno])     
})

## Draw dendrogram with variable values as colors
plotDendroAndColors(obstree2, trait_colors,
                    groupLabels = colnames(trait_colors),
                    dendroLabels = FALSE,
                    cex.dendroLabels = 0.2,
                    main = "Sample dendrogram and trait heatmap")
```


# Network Construction and Module Detection 
## Choose soft threshold 
First, use the `pickSoftThreshold` function to calculate the scale-free
topology fit index and connectivity over a range of candidate soft
threshold values. 

```{r test-soft-thresholds}
powersvec <- c(c(1:10), seq(from = 12, to=20, by=2))
sft <- pickSoftThreshold(protdat, powerVector = powersvec, verbose = 5)

## Extract fit indices
fitindices <- data.table(sft$fitIndices)
```


```{r visualize-soft-thresholds, include=TRUE}
par(mfrow = c(1,2), mar=c(5,4,4,4))
cex1 = 0.9

## Define signed r2 
r2forplot <- -sign(fitindices[, slope])*fitindices[, SFT.R.sq]

## Plot power v. model fit
plot(fitindices[, Power], r2forplot,
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit,signed R^2",
     type = "n",
     main = paste("Scale independence"))
text(fitindices[, Power], r2forplot, labels = powersvec, cex = cex1, col = "red");
abline(h = 0.90, col = "red") # this line corresponds to using an R^2 cut-off of h

## Plot power v. connectivity
plot(fitindices[, Power], fitindices[, mean.k.],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type="n",
     main = paste("Mean connectivity"))
text(fitindices[, Power], fitindices[, mean.k.], labels = powersvec, cex=cex1, col="red")

```


The WGCNA tutorial recommends choosing the lowest soft threshold value
after the scale-free topology model fit levels off, as long as it is
above 0.8. The red horizontal line indicates r2 of 0.9. 
Here, based on the plot, I chose a value of 
`r args$softpower`.

## Use one-step network construction and module detection function 
The easiest way to construct the network is to use the one-step
function. In addition to the parameter `power` for which we have
chosen the value of `r args$softpower` based on the plots
above, We can set values for some other potentially important parameters 
for the network construction -- see function call below for all values. 

```{r one-step-construction, results=FALSE, echo=TRUE}
net <- blockwiseModules(protdat,
                        power = args$softpower,
                        corType = "bicor",
                        networkType = "signed",
                        TOMType = "signed",
                        maxBlockSize = 10000,
                        maxPOutliers = 0.10, 
                        minModuleSize = 20,
                        reassignThreshold = 0,
                        deepSplit = 4,
                        mergeCutHeight = args$cutheight,
                        numericLabels = FALSE,
                        pamRespectsDendro = FALSE,
                        saveTOMs = TRUE,
                        saveTOMFileBase = outfiles$tombase,
                        verbose = 3)

## Extract adjacency matrix
adj_mat <- adjacency(protdat,
                     power = args$softpower,
                     corFnc = "bicor",
                     type = "signed",                     
                     corOptions = list(use = 'pairwise.complete.obs'))

```


# Diagnostics

```{r get-module-colors}
moduleColors <- net$colors
nmodules <- length(unique(moduleColors))
```


The automatic network construction identified 
`r nmodules-1` modules, not including the grey 'module'. 

## Network dendrogram 

```{r visualize-network, include=TRUE}
plotDendroAndColors(dendro = net$dendrograms[[1]],
                    colors = moduleColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```


## Number of proteins per module 

```{r visualize-module-sizes-barplot, include=TRUE}
## Draw barplot of the number of proteins per module
par(mar = c(5,5,5,5))
toplot <- sort(table(moduleColors), decreasing=TRUE)
toplot <- toplot[c("grey", names(toplot)[!names(toplot) %in% "grey"])]
colors.ordered <- names(toplot)

toplot.df <- data.frame(module = names(toplot),
                        value = as.numeric(toplot),
                        modcolor = colors.ordered)
toplot.df$module <- factor(toplot.df$module, levels = toplot.df$module)


ggplot(toplot.df, aes(x = module, y = value, fill = modcolor)) +
    geom_bar(stat = "identity") +
    scale_fill_identity() +
    geom_text(aes(label = value), vjust = -0.3, size = 3) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    labs(x = "Module", y = "Number of proteins", title = "Protein count per module")
```


## Module membership
There are few different metrics for characterizing proteins. 

* kME -- also known as module membership, this is the correlation
  between protein and the module eigengene. We can
  calculate kME between the protein and any
  module eigengene, but we probably want to look at it for the module to which
  the protein has been assigned. 

* kIM -- also known as intramodule connectivity, this is the sum of the adjacencies to
  the other proteins in the same module. 
  
* soft connectivity -- the sum of the adjacencies to all other nodes 

```{r calculate-eigengenes, include=FALSE}
## Calculate eigengene for each modulue
MEList <- moduleEigengenes(protdat, colors = moduleColors)
MEs0 <- MEList$eigengenes
MEs <- orderMEs(MEs0)
rm(list = c("MEList", "MEs0")) # remove tmp objects

eigengenes_forprint <- MEs
eigengenes_forprint$protsample <- rownames(eigengenes_forprint)
setDT(eigengenes_forprint)
setcolorder(eigengenes_forprint, "protsample")
```


```{r calculate-connectivities, include=FALSE}
### Calculate kME for each protein to every eigengene
kMEs <- data.table(protein = colnames(protdat),
                   signedKME(datExpr = protdat,
                             datME = MEs))

## Calculate connectivities -- whole network, within module, outside module
## From adjacency matrix
intramod_conn.adj <- data.table(
    protein = colnames(protdat),
    intramodularConnectivity(
    adj_mat,
    colors = moduleColors
    )
)

## Make merged table with connectivity stats and assigned module
conn_stats <- merge(data.table(protein = names(moduleColors), module = moduleColors),
                    intramod_conn.adj,                        
                    by = "protein")

## Add the kME of each protein to its assigned module
for(amod in colors.ordered) {
    for(aprotein in conn_stats[module == amod, protein]) {
        conn_stats[protein == aprotein, kME := kMEs[protein == aprotein, paste0("kME", amod),
                                                    with = FALSE]]
    }
}
```


Here is the distribution of kMEs per module. 
```{r visualize-kME, eval=TRUE, fig.height=10, fig.width = 9, eval = FALSE}
par(mfrow=c(ceiling(nmodules/4),4)) # define number of rows based on number of modules
for(acolor in c(setdiff(colors.ordered, "grey"), "grey")) { ## plot grey module last
    hist(conn_stats[module == acolor, kME],
         col=acolor, xlim=c(-1,1), main=acolor, xlab="")
}
```



```{r visualize-kIM, include=TRUE, eval = FALSE}
## Then we can look at the intramodule connectivities per module,
## but these should probably be scaled because proteins in larger modules
## will have larger values since it is a sum
par(mfrow=c(ceiling(nmodules/4),4), oma=c(1,1,3,1)) # define num rows based on number of modules
for(acolor in c(setdiff(colors.ordered, "grey"), "grey")) { ## plot grey module last
    axlim=c(0, max(conn_stats[,"kWithin"], na.rm=TRUE))
    tmptoplot <- conn_stats[module==acolor, kWithin]
    hist(tmptoplot, xlim=axlim, col=acolor, main=acolor, xlab="within module connectivity")
    rm(list=c("tmptoplot", "axlim"))
}

mtext("Distribution of intramodule connectivity per module", outer=TRUE)
```


# Module association with individual NPS 
Here we test association between each module eigengene and our traits
of interest. 

```{r run-regressions, include=FALSE}
### Set up object to hold regression results
allsumms <- vector(mode="list")
save.regression.results <- FALSE

### Iterate over outcomes
for(anps in nps) {
    aform <- paste0(anps, "~ mod.eigen")
    atype <- "binomial"

    ## run regression for each module (exclude grey)
    regression.results <- lapply(setdiff(colnames(MEs), "grey"), function(amodule) {
        regression_data <- PrepareDataME(amodule, phenos, MEs)
        res.onemodule <- glm(form = aform, data = regression_data, family = "binomial")
        return(res.onemodule)
    })
    
    ## Summarize results for this outcome
    regression.summ <- data.frame(
        coeff.stderr = t(sapply(regression.results,
                                GetCoefficientAndSE, type=atype, var="mod.eigen")),
        pvalues = sapply(regression.results, GetPvalue, type=atype, var="mod.eigen"))
    colnames(regression.summ) <- c("Estimate", "Std.Err", "Pvalue")
    rownames(regression.summ) <- colnames(MEs)

    ## Clean up
    print(paste(" model", anps, "--", "Cleaning up"))
    allsumms[[anps]] <- regression.summ
    rm(list=c("regression.results", "regression.summ", "aform", "atype"))
}

names(allsumms) <- nps

### Prepare results for matrix visualization
coeffs <- Reduce(cbind, lapply(allsumms, function(x) {signif(x$Estimate, 2) }))
pvals <- Reduce(cbind, lapply(allsumms, function(x) {signif(x$Pvalue, 2) }))
rownames(pvals) <- rownames(coeffs) <- colnames(MEs)
colnames(pvals) <- colnames(coeffs) <- names(allsumms)

textMatrix.regression <- paste0(coeffs, "\n(",
                                pvals,  ")")
dim(textMatrix.regression) <- dim(coeffs)

## Get index for smallest p-value
minind <- which(pvals==min(pvals), arr.ind=TRUE)

ntests <- nrow(pvals) * ncol(pvals)
bonfthresh <- 0.05/ntests

```


The heatmap is colored by the regression coefficient. The association
p-value is given in parentheses.

The smallest p-value is `r sprintf("%.1e", min(pvals))`, for the association between 
eigengene `r rownames(pvals)[minind[1]] ` and
`r colnames(pvals)[minind[2]]`. 

The number of tests is `r ntests`, and the Bonferroni threshold at alpha 0.05
is `r sprintf("%.1e", bonfthresh)`.

```{r visualize-association-results, include=TRUE, fig.height=11}
par(mar = c(6, 8.5, 3, 3))
zmax <- max(abs(coeffs))

labeledHeatmap(Matrix = coeffs,
               xLabels = names(allsumms),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = colorRampPalette(c("blue", "white", "red"))(30),
               textMatrix = textMatrix.regression,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(zmax*-1, zmax),
               main = paste("Module-trait relationships (NPS ~ eigengene + sex)"))
```



```{r identify-sig-modules}
### Reformat association p-values to data table with module as column
pvals_dt <- data.frame(pvals)
pvals_dt$module <- sapply(rownames(pvals), function(x) { gsub("ME", "", x) })
setDT(pvals_dt)
setnames(pvals_dt, nps, paste0("pval_", nps))

## Get min p-value and corresponding NPS for each module
pval_cols <- paste0("pval_", nps)
pvals_dt[, `:=` (
    min_p = do.call(pmin, c(.SD, list()))
    ), .SDcols = pval_cols]

### Identify significant module-NPS association at several thresholds
Getsig <- function(pvalrow, sigthresh = 0.05) {
    ## get vector of pval_col names
    vec <- pval_cols[which(pvalrow < sigthresh)] 

    # reformat as comma-delimited list of NPS names
    paste(sapply(vec, function(x) { gsub("pval_", "", x) }), collapse = ", ") 
}

pvals_dt[, NPS_sig05 := apply(.SD, 1, Getsig, sigthresh = 0.05), .SDcols = pval_cols]
pvals_dt[, NPS_sig01 := apply(.SD, 1, Getsig, sigthresh = 0.01), .SDcols = pval_cols]
setcolorder(pvals_dt, c("module", "min_p", "NPS_sig05", "NPS_sig01"))
pvals_dt[, module := factor(module, levels = colors.ordered)]
setorder(pvals_dt, module)
```

Here are the modules that are significant at various thresholds. There
are 
`r sum(pvals_dt[, min_p < 0.05])` and 
`r sum(pvals_dt[, min_p < 0.01])` modules with significant association
at nominal p-values 0.05 and 0.01,
respectively. 


```{r table-sig-modules}
knitr::kable(pvals_dt)
```


```{r extract-hub-genes}
### Extract hub genes
hubgenes <- conn_stats[module != "grey" & kME > 0.7, .(protein, module, kME)]
hubgenes[, module := factor(module, levels = colors.ordered)]
setorder(hubgenes, module, -kME)
hubgenes[, gene_symbol := sapply(protein, function(x) {
    unlist(strsplit(x, split = "\\|"))[1] })]


### Make table of total gene counts and hub gene counts
n_genes <- merge(data.table(table(conn_stats$module))[, .(module = V1, n_total_genes = N)],
                 data.table(table(hubgenes$module))[, .(module = V1, n_hub_genes = N)],
                 by = "module")

n_genes[, module := factor(module, levels = colors.ordered)]
setorder(n_genes, module)

```

```{r make-background-gene-list}
background_genes <- data.table(protein = colnames(protdat))
background_genes[, gene_symbol := sapply(protein, function(x) {
    unlist(strsplit(x, split = "\\|"))[1] })]

## Fill in NA for genes missing gene symbol in protein name
background_genes[gene_symbol == "", gene_symbol := NA]

```


```{r finish}
### Write proteins connectivity stats
write.table(conn_stats, file = outfiles[["conn_stats"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write eigengenes
write.table(eigengenes_forprint, file = outfiles[["eigengenes"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write hub gene stats
write.table(hubgenes, file = outfiles[["hubgenes"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write gene counts
write.table(n_genes, file = outfiles[["n_genes"]], 
            row.names = FALSE, quote = FALSE, sep = "\t")


### Prepare then write regression stats
coeffs.dt <- data.frame(coeffs)
pvals.dt <- data.frame(pvals)
coeffs.dt$module <- rownames(coeffs)
pvals.dt$module <- rownames(pvals)
setDT(coeffs.dt)
setDT(pvals.dt)
setnames(coeffs.dt, nps, paste0(nps, "_coeff"))
setnames(pvals.dt, nps, paste0(nps, "_pval"))

regress_stats <- merge(coeffs.dt, pvals.dt, by = "module")

write.table(regress_stats, file = outfiles[["regress_stats"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write significant module-trait association table
write.table(pvals_dt, file = outfiles[["sigmodules"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write module genes for modules with nominally significant NPS association
invisible(sapply(as.character(pvals_dt[!NPS_sig05 == "", module]), function(amodule) {
    proteins <- conn_stats[module == amodule, protein ]
    genes <- sapply(proteins, function(x) { unlist(strsplit(x, split = "\\|"))[1] })
    write(genes, ncol = 1, file = Outfile_module_genes(amodule))
}))


### Write log
write(paste0("Number of modules significantly associated with NPS"), file = outfiles[["log"]])
write(paste0("     at p-value < 0.05: ",
             sum(pvals_dt[, min_p < 0.05])), file = outfiles[["log"]], append = TRUE)
write(paste0("     at p-value < 0.01: ",
             sum(pvals_dt[, min_p < 0.01])), file = outfiles[["log"]], append = TRUE)

### Write background genes list
write.table(background_genes[, .(gene_symbol)], file = outfiles[["background_genes"]],
            row.names = FALSE,
            quote = FALSE, sep = "\t", col.names = FALSE)

### Write adjacency matrix
write.table(adj_mat, file = outfiles[["adj_df"]],
            row.names = TRUE, quote = FALSE, sep = "\t")

### Write expression data
protdat_forprint <- as.data.frame(protdat)
protdat_forprint$protsample <- rownames(protdat)
setDT(protdat_forprint)
setcolorder(protdat_forprint, "protsample")
fwrite(protdat_forprint, file = outfiles[["protdat"]],
       row.names = FALSE, quote = FALSE, sep = "\t")
```
