---
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: html_document
---


```{r goal, include = FALSE}
###### GOAL ######
```


```{r online-resources, include = FALSE}
## WGCNA package documentation
## https://www.rdocumentation.org/packages/WGCNA/versions/1.73

## one tutorial (there are many online)
## https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0

## hub gene selection
## https://support.bioconductor.org/p/69858/#70116
```


```{r load-packages, include = FALSE}
library(data.table)
library(argparse)
library(WGCNA)
library(ggplot2)
library(robust)

rm(list = ls())
```


```{r functions, include = FALSE}
###### FUNCTIONS ######
source("~/comsv/Svattathil_Library/svattathil_functions.r")

GetCoefficientAndSE <- function(regression.out, type, var="miRNA") {
    if(type=="ordinal") { return(summary(regression.out)$coefficients[var, c("Value", "Std. Error")]) }
    else if(type %in% c("gaussian", "binomial", "lmm")) {
        return(summary(regression.out)$coefficients[var, c("Estimate", "Std. Error")])
    }
}

GetPvalue <- function(regression.out, type, var="miRNA") {
    if(type=="ordinal") {
        return(pnorm(abs(summary(regression.out)$coefficient[var, "t value"]), lower.tail=FALSE) * 2) }
    else if(type %in% c("binomial", "neg.binomial")) { summary(regression.out)$coefficients[var, "Pr(>|z|)"] }
    else if(type %in% c("gaussian", "lmm")) { summary(regression.out)$coefficients[var, "Pr(>|t|)"] }
}

GetStats <- function(regression.out, modelfamily = "binomial", var="protein") {
    ## Extract estimate, standard error, and P-value for protein
    pcol <- ifelse(modelfamily == "binomial", "Pr(>|z|)", "Pr(>|t|)") ## labels differ for different model types
    stats <- c(summary(regression.out)$coefficients[var, c("Estimate", "Std. Error", pcol)])
    ## Add number of samples that were included in analysis
    toreturn <- c(stats, nobs=length(regression.out$residuals) )
    names(toreturn) <- c("Estimate", "SE", "P", "nobs")
    return(toreturn)
}


```


```{r setup, include = FALSE}
###### SETUP ######
### Command line arguments
## Create parser
parser <- ArgumentParser()
parser$add_argument("--cohort", type="character", help="OHSU, rush, or emory", default = "emory")
parser$add_argument("--softpower", type="integer", help="value for softpower", default = 9)
parser$add_argument("--cutheight", type="double", help="value for mergeCutHeight", default = 0.30)
parser$add_argument("--network_set", type="character", help="cohort-specific or consensus",
                    default = "cohort-specific")


## Read from parser
args <- parser$parse_args()


### Constants
options(stringsAsFactors = FALSE)

```


---
title: "Module association with latent factors for `r args$cohort`" 
---


```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```


```{r word-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```


```{r input}
### Files that exist
datdir <- "2_Pipeline/02-Prepare_analysis_data/"

files <- list()
files[["phenos"]]        <- paste0(datdir, "phenos_cleaned_SVs_", args$cohort, ".txt")
files[["factor_scores"]] <- paste0(datdir, "Latent_factors/factor_scores_", args$cohort, ".txt")

if(args$network_set == "cohort-specific") {
    indir <- "2_Pipeline/04-Build_networks/Cohort-specific/Selected_results/"
    runid <- paste0(args$cohort, "_softpower", args$softpower,
                    "_cutheight", sprintf("%.2f", args$cutheight))
}

if(args$network_set == "consensus") {
    indir <- paste0("2_Pipeline/04-Build_networks/Wgcna_out/Consensus/Auto_softpower", args$softpower, "/")
    runid <- paste0("auto_softpower", args$softpower, "_", args$cohort)
    
}

files[["eigengenes"]] = paste0(indir, runid, "_eigengenes.txt")

```


```{r output}
### Files to be created
outdir <- paste0(indir, "DEA_latent_factors/")
MyMkdir(outdir)

outfiles <- lapply(list(
    regress_stats    = "_eigengene_regress_stats_latent_factors.txt",
    sigmodules       = "_associated_modules_latent_factors.txt"
), function(x) { paste0(outdir, runid, x) })


```


```{r read-data}
###### MAIN ######
### Read in data
eigengenes <- fread(files[["eigengenes"]])
phenos.all <- fread(files[["phenos"]])
factor_scores <- fread(files[["factor_scores"]])
                    
```


```{r prepare-data}
### Prepare data
### Define factor names
factor_names <- setdiff(names(factor_scores), "protsample")


### Define module eigengenes of interest (exclude grey)
mes_to_test <- setdiff(colnames(eigengenes), c("protsample", "MEgrey"))


### Merge phenos and factor scores
phenos <- merge(phenos.all[, .(protsample, msex)], factor_scores, by = "protsample")


### Add binary versions of factors
for(afactor in factor_names) {
    currfactor <- unlist(phenos[, ..afactor])
    phenos[, paste0(afactor, ".2lev") := ifelse(currfactor < median(currfactor), 0, 1)]
}
rm(afactor)



```

## Distribution of factor scores

For the `r args$cohort` dataset, we constructed 
`r ncol(factor_scores)-1` latent factors based on the individual
binary NPS variables (e.g., agitation, apathy anxiety...). 

The vertical dotted line indicates the median.

```{r visualize-score-distributions, fig.height = 8}
nfactors <- length(factor_names)
par(mfrow=c(nfactors, 1))
invisible(sapply(factor_names, function(afactor) {
    toplot <- unlist(factor_scores[, afactor, with = FALSE])
    hist(toplot, main = afactor, xlab = "", breaks = 30)
    abline(v = median(toplot), lty = 2)
}))
```

## Distribution of module eigengenes
The WGCNA analysis identified `r length(mes_to_test)` modules. Here is the distribution of
the eigengene for each module. 

 
```{r visualize-eigengene-distributions, fig.height = 8}
nmodules <- length(mes_to_test)
par(mfrow=c(ceiling(nmodules/4),4)) # define number of rows based on number of modules

invisible(sapply(mes_to_test, function(amodule) {
    acol <- gsub("ME", "", amodule)
    toplot <- unlist(eigengenes[, ..amodule])
    hist(toplot, main = amodule, xlab = "", breaks = 30,
         col = acol, border = "black", lwd=0.5)
}))

```


## Module association with latent factors as continuous outcomes
Here we test association between each module eigengene and latent
factors constructed based on the individual NPS.


```{r run-regressions-gaussian, include=FALSE}
## Define outcome distribution type
vartype <- "gaussian"

## Set up object to hold regression results
## summs_list will be a list of data frames --
##   each dataframe gives regression summary statistics, one row per module
## resobj_list will be a list of lists --
##   each sublist is a list of the fit objects for each module
summs_list <- vector(mode="list")
resobj_list <- vector(mode="list")


### Iterate over outcomes
for(afactor in factor_names) {
    aform <- paste0("mod.eigen ~ ", afactor)
    
    ## Run regression for each module eigengene (exclude grey)
    
    regression.results <- lapply(mes_to_test, function(amodule) {
        ## Prepare regression data
        ## Combine eigengenes with other necessary phenos
        regression_data <- merge(eigengenes[, c("protsample", amodule), with = FALSE], phenos, by = "protsample")
        setnames(regression_data, amodule, "mod.eigen") ## rename eigengene column to generic name
        
        ## Convert to data frame and add protsample as rownames
        samps <- regression_data$protsample
        setDF(regression_data)
        rownames(regression_data) <- samps

        ## Run regression
        res.onemodule <- lm(form = aform, data = regression_data, na.action = "na.exclude")
        return(res.onemodule)
    })
    names(regression.results) <- mes_to_test

    
    ## Summarize results for this outcome
    regression.summ <- data.frame(t(sapply(regression.results, GetStats,
                                           modelfamily = vartype, var = afactor)))
    rownames(regression.summ) <- mes_to_test

    ## Clean up
    print(paste(" model", afactor, "--", "Cleaning up"))
    summs_list[[afactor]] <- regression.summ
    resobj_list[[afactor]] <- regression.results
    rm(list=c("regression.results", "regression.summ", "aform"))
}
rm(afactor)

names(summs_list) <- factor_names
```

```{r prepare-matrix-vis-gaussian}
### Prepare results for matrix visualization
## Use cbind to ensure that object is a matrix even when there is only one latent factor
coeffs <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$Estimate, 2) })))
pvals <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$P, 2) })))
rownames(pvals) <- rownames(coeffs) <- mes_to_test
colnames(pvals) <- colnames(coeffs) <- names(summs_list)

textMatrix.regression <- paste0(coeffs, "\n(",
                                pvals,  ")")
dim(textMatrix.regression) <- dim(coeffs)

## Get index for smallest p-value
minind <- which(pvals==min(pvals), arr.ind=TRUE)

ntests <- nrow(pvals) * ncol(pvals)
bonfthresh <- 0.05/ntests
```


The heatmap is colored by the regression coefficient. The association
p-value is given in parentheses.

The smallest p-value is `r sprintf("%.1e", min(pvals))`, for the association between 
eigengene `r rownames(pvals)[minind[1]] ` and
`r colnames(pvals)[minind[2]]`. 

The number of tests is `r ntests`, and the Bonferroni threshold at alpha 0.05
is `r sprintf("%.1e", bonfthresh)`.

```{r visualize-association-results-gaussian, include=TRUE, fig.height=11}
par(mar = c(6, 8.5, 3, 3))
zmax <- max(abs(coeffs))

labeledHeatmap(Matrix = coeffs,
               xLabels = names(summs_list),
               yLabels = rownames(coeffs),
               ySymbols = rownames(coeffs),
               yColorLabels = TRUE,
               colors = colorRampPalette(c("blue", "white", "red"))(30),
               textMatrix = textMatrix.regression,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(zmax*-1, zmax),
               main = paste("Module-trait relationships (latentfactor ~ eigengene + sex)"))
```


```{r identify-sig-modules-gaussian}
### Reformat association p-values to data table with module as column
pvals_dt <- data.frame(pvals)
pvals_dt$module <- sapply(rownames(pvals), function(x) { gsub("ME", "", x) })
setDT(pvals_dt)
setnames(pvals_dt, factor_names, paste0("pval_", factor_names))

## Get min p-value and corresponding latentfactor for each module
pval_cols <- paste0("pval_", factor_names)
pvals_dt[, `:=` (
    min_p = do.call(pmin, c(.SD, list()))
    ), .SDcols = pval_cols]

### Identify significant module-NPS association at several thresholds
Getsig <- function(pvalrow, sigthresh = 0.05) {
    ## get vector of pval_col names
    vec <- pval_cols[which(pvalrow < sigthresh)] 

    # reformat as comma-delimited list of NPS names
    paste(sapply(vec, function(x) { gsub("pval_", "", x) }), collapse = ", ") 
}

pvals_dt[, latentfactor_sig05 := apply(.SD, 1, Getsig, sigthresh = 0.05), .SDcols = pval_cols]
pvals_dt[, latentfactor_sig01 := apply(.SD, 1, Getsig, sigthresh = 0.01), .SDcols = pval_cols]
setcolorder(pvals_dt, c("module", "min_p", "latentfactor_sig05", "latentfactor_sig01"))
```

Here are the modules that are significant at various thresholds. There
are 
`r sum(pvals_dt[, min_p < 0.05])` and 
`r sum(pvals_dt[, min_p < 0.01])` modules with significant association
at nominal p-values 0.05 and 0.01,
respectively. 


```{r table-sig-modules-gaussian}
knitr::kable(pvals_dt)
```

## Check residual plots to evaluate model fit
Here are the plots of residuals v. fitted values and the distribution
of residuals for any associations
significant at unadjusted p-value < 0.05.


```{r check-residuals-gaussian}
### Identify nominally significant associations
sigassoc <- which(pvals < 0.05, arr.ind =TRUE)
if(nrow(sigassoc) > 0) {
    for(i in 1:nrow(sigassoc)) {
        par(oma = c(1, 1, 1.5, 1))
        curr_me <- rownames(pvals)[sigassoc[i, "row"]]
        curr_factor <- colnames(pvals)[sigassoc[i, "col"]]
        curr_res <- resobj_list[[curr_factor]][[curr_me]]


        par(mfrow=c(1, 2))
        plot(fitted(curr_res), residuals(curr_res),
             main = "Residuals v. fitted values",
             xlab = "Fitted Values", 
             ylab = "Residuals")
        abline(h = 0, col = "red", lty = 2)

        plot(density(residuals(curr_res)),
             main = "Density of residuals")
        mtext(paste0("Factor ", curr_factor, " association with ", curr_me), outer = TRUE)    
    }
}
```


```{r finish-gaussian}
### Prepare then write regression stats
coeffs.dt <- data.frame(coeffs)
pvals.dt <- data.frame(pvals)
coeffs.dt$module <- rownames(coeffs)
pvals.dt$module <- rownames(pvals)
setDT(coeffs.dt)
setDT(pvals.dt)
setnames(coeffs.dt, factor_names, paste0(factor_names, "_coeff"))
setnames(pvals.dt, factor_names, paste0(factor_names, "_pval"))

regress_stats <- merge(coeffs.dt, pvals.dt, by = "module")

### Write table of regression summary statistics
write.table(regress_stats, file = outfiles[["regress_stats"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write significant module-trait association table
write.table(pvals_dt, file = outfiles[["sigmodules"]],
            row.names = FALSE, quote = FALSE, sep = "\t")

```


## Module association with latent factors as binary outcomes
Since the latent factors tend to be almost bimodal and the
residuals v. fitted plots show that residuals are not randomly
distributed, maybe the model is not very reliable. So, I also
transformed the latent factors to binary outcomes
using median dichotomization and then used logistic regression to test
for module association. 


```{r run-regressions-binary, include=FALSE, eval = TRUE}
## Define outcome distribution type
vartype <- "binomial"

## Set up object to hold regression results
## summs_list will be a list of data frames --
##   each dataframe gives regression summary statistics, one row per module
## resobj_list will be a list of lists --
##   each sublist is a list of the fit objects for each module
summs_list <- vector(mode="list")
resobj_list <- vector(mode="list")


### Iterate over outcomes
for(afactor in factor_names) {
    ## Define model form
    aform <- paste0(paste0(afactor, ".2lev"), " ~ mod.eigen + msex")
    
    ## Set seed since lmRob produces approximations by a random algorithm
    robust::lmRob.control(seed=523832, mxf=200, mxr=200, mxs=200)
    
    ## Run regression for each module eigengene (exclude grey)
    
    regression.results <- lapply(mes_to_test, function(amodule) {
        ## Prepare regression data
        ## Combine eigengenes with other necessary phenos
        regression_data <- merge(eigengenes[, c("protsample", amodule), with = FALSE], phenos, by = "protsample")
        setnames(regression_data, amodule, "mod.eigen") ## rename eigengene column to generic name
        
        ## Convert to data frame and add protsample as rownames
        samps <- regression_data$protsample
        setDF(regression_data)
        rownames(regression_data) <- samps

        ## Run regression
        res.onemodule <- robust::glmRob(form = aform, data = regression_data,
                                        family = "binomial", na.action = "na.exclude")
        return(res.onemodule)
    })
    names(regression.results) <- mes_to_test

    
    ## Summarize results for this outcome
    regression.summ <- data.frame(t(sapply(regression.results, GetStats, modelfamily = vartype, var = "mod.eigen")))
    rownames(regression.summ) <- mes_to_test

    ## Clean up
    print(paste(" model", afactor, ".2lev --", "Cleaning up"))
    summs_list[[afactor]] <- regression.summ
    resobj_list[[afactor]] <- regression.results
    rm(list=c("regression.results", "regression.summ", "aform"))
}
rm(afactor)

names(summs_list) <- factor_names
```


```{r prepare-matrix-vis-binary}
### Prepare results for matrix visualization
## Use cbind to ensure that object is a matrix even when there is only one latent factor
coeffs <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$Estimate, 2) })))
pvals <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$P, 2) })))
rownames(pvals) <- rownames(coeffs) <- mes_to_test
colnames(pvals) <- colnames(coeffs) <- names(summs_list)

textMatrix.regression <- paste0(coeffs, "\n(",
                                pvals,  ")")
dim(textMatrix.regression) <- dim(coeffs)

## Get index for smallest p-value
minind <- which(pvals==min(pvals), arr.ind=TRUE)

ntests <- nrow(pvals) * ncol(pvals)
bonfthresh <- 0.05/ntests
```

The smallest p-value is `r sprintf("%.1e", min(pvals))`, for the association between 
eigengene `r rownames(pvals)[minind[1]] ` and
`r colnames(pvals)[minind[2]]`. 

The number of tests is `r ntests`, and the Bonferroni threshold at alpha 0.05
is `r sprintf("%.1e", bonfthresh)`.

```{r visualize-association-results-binary, include=TRUE, fig.height=11}
par(mar = c(6, 8.5, 3, 3))
zmax <- max(abs(coeffs))

labeledHeatmap(Matrix = coeffs,
               xLabels = names(summs_list),
               yLabels = rownames(coeffs),
               ySymbols = rownames(coeffs),
               yColorLabels = TRUE,
               colors = colorRampPalette(c("blue", "white", "red"))(30),
               textMatrix = textMatrix.regression,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(zmax*-1, zmax),
               main = paste("Module-trait relationships (latentfactor (binary) ~ eigengene + sex)"))
```


```{r identify-sig-modules-binary}
### Reformat association p-values to data table with module as column
pvals_dt <- data.frame(pvals)
pvals_dt$module <- sapply(rownames(pvals), function(x) { gsub("ME", "", x) })
setDT(pvals_dt)
setnames(pvals_dt, factor_names, paste0("pval_", factor_names))

## Get min p-value and corresponding latentfactor for each module
pval_cols <- paste0("pval_", factor_names)
pvals_dt[, `:=` (
    min_p = do.call(pmin, c(.SD, list()))
    ), .SDcols = pval_cols]

### Identify significant module-NPS association at several thresholds
Getsig <- function(pvalrow, sigthresh = 0.05) {
    ## get vector of pval_col names
    vec <- pval_cols[which(pvalrow < sigthresh)] 

    # reformat as comma-delimited list of NPS names
    paste(sapply(vec, function(x) { gsub("pval_", "", x) }), collapse = ", ") 
}

pvals_dt[, latentfactor_sig05 := apply(.SD, 1, Getsig, sigthresh = 0.05), .SDcols = pval_cols]
pvals_dt[, latentfactor_sig01 := apply(.SD, 1, Getsig, sigthresh = 0.01), .SDcols = pval_cols]
setcolorder(pvals_dt, c("module", "min_p", "latentfactor_sig05", "latentfactor_sig01"))
```

Here are the modules that are significant at various thresholds. There
are 
`r sum(pvals_dt[, min_p < 0.05])` and 
`r sum(pvals_dt[, min_p < 0.01])` modules with significant association
at nominal p-values 0.05 and 0.01,
respectively. 


```{r table-sig-modules-binary}
knitr::kable(pvals_dt)
```


## Check residual plots to evaluate model fit
These plots would be more useful if I added the loess line to check
linearity. If we want to use these plots, I should add that. 

```{r check-residuals-binary, fig.width = 4}
### Identify nominally significant associations
sigassoc <- which(pvals < 0.05, arr.ind =TRUE)
if(nrow(sigassoc) > 0) {
    for(i in 1:nrow(sigassoc)) {
        par(oma = c(1, 1, 1.5, 1))
        curr_me <- rownames(pvals)[sigassoc[i, "row"]]
        curr_factor <- colnames(pvals)[sigassoc[i, "col"]]
        curr_res <- resobj_list[[curr_factor]][[curr_me]]


        par(mfrow=c(1, 1))
        plot(fitted(curr_res), residuals(curr_res, type = "deviance"),
             main = "Deviance residuals v. fitted values",
             xlab = "Fitted Values", 
             ylab = "Deviance Residuals")
        abline(h = 0, col = "red", lty = 2)

        mtext(paste0("Factor ", curr_factor, " association with ", curr_me), outer = TRUE)   
    }
}
```


## Module association with eigengenes as outcomes
Lastly, since the module eigengenes look more normally distributed
than the latent factors, I tried ordinary linear regression using the
module eigengenes as the outcomes and the latent factors as the predictors. 



```{r run-ordinary-gaussian, include=FALSE}
## Define outcome distribution type
vartype <- "gaussian"

## Set up object to hold regression results
## summs_list will be a list of data frames --
##   each dataframe gives regression summary statistics, one row per module
## resobj_list will be a list of lists --
##   each sublist is a list of the fit objects for each module
summs_list <- vector(mode="list")
resobj_list <- vector(mode="list")


### Iterate over latent factors
for(afactor in factor_names) {
    aform <- paste0("mod.eigen", " ~ ", afactor, "+ msex")
    
    ## Run regression for each module eigengene (exclude grey)
    
    regression.results <- lapply(mes_to_test, function(amodule) {
        ## Prepare regression data
        ## Combine eigengenes with other necessary phenos
        regression_data <- merge(eigengenes[, c("protsample", amodule), with = FALSE], phenos, by = "protsample")
        setnames(regression_data, amodule, "mod.eigen") ## rename eigengene column to generic name
        
        ## Convert to data frame and add protsample as rownames
        samps <- regression_data$protsample
        setDF(regression_data)
        rownames(regression_data) <- samps

        ## Run regression
        res.onemodule <- glm(form = aform, data = regression_data, na.action = "na.exclude", family = vartype)
        return(res.onemodule)
    })
    names(regression.results) <- mes_to_test

    
    ## Summarize results for this outcome
    regression.summ <- data.frame(t(sapply(regression.results, GetStats, modelfamily = vartype, var = afactor)))
    rownames(regression.summ) <- mes_to_test

    ## Clean up
    print(paste(" model", afactor, "--", "Cleaning up"))
    summs_list[[afactor]] <- regression.summ
    resobj_list[[afactor]] <- regression.results
    rm(list=c("regression.results", "regression.summ", "aform"))
}
rm(afactor)

names(summs_list) <- factor_names
```


```{r prepare-matrix-vis-ordinary}
### Prepare results for matrix visualization
## Use cbind to ensure that object is a matrix even when there is only one latent factor
coeffs <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$Estimate, 2) })))
pvals <- cbind(Reduce(cbind, lapply(summs_list, function(x) {signif(x$P, 2) })))
rownames(pvals) <- rownames(coeffs) <- mes_to_test
colnames(pvals) <- colnames(coeffs) <- names(summs_list)

textMatrix.regression <- paste0(coeffs, "\n(",
                                pvals,  ")")
dim(textMatrix.regression) <- dim(coeffs)

## Get index for smallest p-value
minind <- which(pvals==min(pvals), arr.ind=TRUE)

ntests <- nrow(pvals) * ncol(pvals)
bonfthresh <- 0.05/ntests
```

The association p-value is given in parentheses.

The smallest p-value is `r sprintf("%.1e", min(pvals))`, for the association between 
eigengene `r rownames(pvals)[minind[1]] ` and
`r colnames(pvals)[minind[2]]`. 

The number of tests is `r ntests`, and the Bonferroni threshold at alpha 0.05
is `r sprintf("%.1e", bonfthresh)`.

```{r visualize-association-results-ordinary, include=TRUE, fig.height=11}
par(mar = c(6, 8.5, 3, 3))
zmax <- max(abs(coeffs))

labeledHeatmap(Matrix = coeffs,
               xLabels = names(summs_list),
               yLabels = rownames(coeffs),
               ySymbols = rownames(coeffs),
               yColorLabels = TRUE,
               colors = colorRampPalette(c("blue", "white", "red"))(30),
               textMatrix = textMatrix.regression,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(zmax*-1, zmax),
               main = paste("Module-trait relationships (eigengene ~ latentfactor + sex)"))
```

```{r identify-sig-modules-ordinary}
### Reformat association p-values to data table with module as column
pvals_dt <- data.frame(pvals)
pvals_dt$module <- sapply(rownames(pvals), function(x) { gsub("ME", "", x) })
setDT(pvals_dt)
setnames(pvals_dt, factor_names, paste0("pval_", factor_names))

## Get min p-value and corresponding latentfactor for each module
pval_cols <- paste0("pval_", factor_names)
pvals_dt[, `:=` (
    min_p = do.call(pmin, c(.SD, list()))
    ), .SDcols = pval_cols]

### Identify significant module-NPS association at several thresholds
Getsig <- function(pvalrow, sigthresh = 0.05) {
    ## get vector of pval_col names
    vec <- pval_cols[which(pvalrow < sigthresh)] 

    # reformat as comma-delimited list of NPS names
    paste(sapply(vec, function(x) { gsub("pval_", "", x) }), collapse = ", ") 
}

pvals_dt[, latentfactor_sig05 := apply(.SD, 1, Getsig, sigthresh = 0.05), .SDcols = pval_cols]
pvals_dt[, latentfactor_sig01 := apply(.SD, 1, Getsig, sigthresh = 0.01), .SDcols = pval_cols]
setcolorder(pvals_dt, c("module", "min_p", "latentfactor_sig05", "latentfactor_sig01"))
```

Here are the modules that are significant at various thresholds. There
are 
`r sum(pvals_dt[, min_p < 0.05])` and 
`r sum(pvals_dt[, min_p < 0.01])` modules with significant association
at nominal p-values 0.05 and 0.01,
respectively. 


```{r table-sig-modules-ordinary}
knitr::kable(pvals_dt)
```

## Check residual plots to evaluate model fit
Here are the plots of residuals v. fitted values and the distribution
of residuals for any associations
significant at unadjusted p-value < 0.05.


```{r check-residuals-ordinary}
### Identify nominally significant associations
sigassoc <- which(pvals < 0.05, arr.ind =TRUE)
if(nrow(sigassoc) > 0) {
    for(i in 1:nrow(sigassoc)) {
        par(oma = c(1, 1, 1.5, 1))
        curr_me <- rownames(pvals)[sigassoc[i, "row"]]
        curr_factor <- colnames(pvals)[sigassoc[i, "col"]]
        curr_res <- resobj_list[[curr_factor]][[curr_me]]


        par(mfrow=c(1, 2))
        plot(fitted(curr_res), residuals(curr_res),
             main = "Residuals v. fitted values",
             xlab = "Fitted Values", 
             ylab = "Residuals")
        abline(h = 0, col = "red", lty = 2)

        plot(density(residuals(curr_res)),
             main = "Density of residuals")
        mtext(paste0("Factor ", curr_factor, " association with ", curr_me), outer = TRUE)    
    }
}
```

```{r finish-ordinary}
### Prepare then write regression stats
coeffs.dt <- data.frame(coeffs)
pvals.dt <- data.frame(pvals)
coeffs.dt$module <- rownames(coeffs)
pvals.dt$module <- rownames(pvals)
setDT(coeffs.dt)
setDT(pvals.dt)
setnames(coeffs.dt, factor_names, paste0(factor_names, "_coeff"))
setnames(pvals.dt, factor_names, paste0(factor_names, "_pval"))

regress_stats <- merge(coeffs.dt, pvals.dt, by = "module")

### Write table of regression summary statistics
write.table(regress_stats, file = outfiles[["regress_stats"]],
            row.names = FALSE, quote = FALSE, sep = "\t")


### Write significant module-trait association table
write.table(pvals_dt, file = outfiles[["sigmodules"]],
            row.names = FALSE, quote = FALSE, sep = "\t")

```

