---
title: "Explore proteomics data"
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
#library(gplots)
library(pheatmap)
library(ggplot2)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }
```

```{r constants}
ids.bad.from.kate <- matrix(c("sample mixture", "E12-06",
                              "inbred", "E13-10",
                              "inbred", "29832086",
                              "bad experiment", "1907"), byrow=TRUE, ncol=2)

ids.dups.from.kate <- matrix(c("1993", "1991",
                               "2563", "2363"), byrow=TRUE, ncol=2)
                          
```


```{r define-input}
### Files that exist
filedir <- "0_Data/SharepointFiles/"
file.prot <- paste0(filedir, "Proteomics/QC/protein_compiled_norm.csv")
file.ids <- paste0(filedir, "Proteomics/QC/traits_assembled.csv")
file.tampor <- paste0(filedir, "Proteomics/QC/3.Regressed.Batch_cleanDat-9284x389-WINGO_BRAIN-TAMPOR+BatchPMIRegression.csv")
```


```{r define-output}
### Files to be created
outdir <- "2_Pipeline/01-Explore_data/"
MyMkdir(outdir)
out.ids.filtered <- paste0(outdir, "ids_semi-filtered.txt")
out.heatmap <- paste0(outdir, "sample_heatmap.png")
out.log2norm <- paste0(outdir, "prot_before_sample_outlier_removal_log2norm.txt")
```


```{r read-data}
### Read in data
prot <- fread(file.prot)
ids0 <- fread(file.ids)
tampor <- fread(file.tampor)
```


```{r prepare-data-1}
## The prot table has first column for protein names and remaining columns for samples

## Update column names
setnames(ids0, "V1", "protsample")


## Remove redundant column
ids0[, ID := NULL]

## set batch as factor
ids0[, Batch := factor(Batch, sort(unique(Batch)))]

```

## Goal
The goal is to explore the proteomics data with regard to data missingness
and batch effects. 

## Data
The data file is located in Sharepoint in this directory:  

Common - Dry Lab > Data_brain proteomics_NPS > QC

The data is in file protein_compiled_norm.csv, and the cohort and a
few other covariates are in file traits_assembled.csv. 



## Results
```{r check-orig-counts, include = FALSE}
## Number of samples
ncol(prot) -1

## Number of proteins
nrow(prot)

## Check how many people from ids are in prot table
table(ids0$protsample %in% colnames(prot), useNA = "ifany")

## Extract ids for samples in prot table
ids.incgis <- copy(ids0[protsample %in% colnames(prot), ])
n.gis <- nrow(ids.incgis[WorkingID == "GIS", ])

## Check how many samples are GIS
ids <- ids.incgis[WorkingID != "GIS", ]

## Select study samples (i.e. exclude GIS)
prot.study <- prot[, c("V1", ids$protsample), with = FALSE]
n.study  <- nrow(ids) ## number of study samples

## Check whether protein has complete NAs
all.missing <- table(apply(prot.study[, !c("V1")], 1, function(x) {all(is.na(x) | x == 0)}))
```
The data file has one column with protein names and remaining columns
for samples. 

There are `r pN(nrow(prot))` proteins in the data file, but some have
complete NAs so there are really `r pN(all.missing["FALSE"])` proteins to work with. We'll be filtering these
down later.   

There are `r ncol(prot) -1` samples in the data file. These include 
`r n.gis` GIS samples and `r nrow(ids)` study samples. 


Here is the count of samples from each cohort: 

```{r table-sources}
## Check how many samples from each source/cohort
Mkable(rbind(sort(table(ids$source, useNA = "ifany"), decreasing=TRUE)))
cohorts <- unique(ids$source)
```


There are `r length(unique(ids$Batch))` batches. Here is a breakdown
of the samples by batch and source/cohort. 

```{r table-batches}
## Count batches and tabulate number of samples in each batch
#length(unique(ids$batch))
kable(table(ids[, .(Batch, source)]), row.names=TRUE, caption = "Samples per batch and source.")

```

### Filtering
#### Filter proteins by minimum sample count
There are both NAs and 0 values in the data. 
Out of curiosity, I visualized the zeros and NAs by protein and
sample. 



```{r make-visual-matrix-zeros-NAs}
### Visualize 0s and NAs
## Convert to matrix
mat <- as.matrix(prot.study[, -c("V1")])

## Create a categorical matrix to classify cells
## Define colors
visual_matrix <- ifelse(is.na(mat), 1, ifelse(mat == 0, 2, 3))
colors <- c("blue", "red", "white")  # Blue for NA, Red for 0, White for other values
```

```{r visualize-zeros-NAs, fig.width=7, fig.height=7, eval=FALSE}
## Draw heatmap with clustering
p1 <- pheatmap(
  visual_matrix,                               # Matrix to be visualized
  cluster_rows = TRUE,                         # Cluster rows
  cluster_cols = TRUE,                         # Cluster columns
  color = colors,                              # Custom colors for 1, 2, 3
  annotation_legend = TRUE,                     # Show legend
  main = "Visualization of 0s and NAs",  # Title
  legend = TRUE,                               # Display the legend
  show_rownames = FALSE,                        # Show row names (proteins)
  show_colnames = TRUE,                        # Show column names (samples)
  labels_row = prot.study$V1,                  # Label proteins
  labels_col = sapply(colnames(mat), substring, 7, 9),  ## label samples only with batch (extract from sample name) 
  fontsize_row = 8,                            # Adjust row label size
  fontsize_col = 5,                            # Adjust column label size
  legend_labels = c("NA", "Zero", "Other"),    # Legend labels
  legend_breaks = c(1:3), 
  )

p1
```

```{r draw-plot-png, eval=FALSE}
png(file = out.heatmap, width = 20, height = 7, res = 300, units = "in")
p1
dev.off()

```

The samples apparently cluster by batch. There are a few proteins with
lots of zeros. There is a set of six batches (leftmost chunk of plot)
with more zeros and NAs than the rest of the samples. 


### Determine number of proteins meeting range of missingness thresholds
For analysis, convert the 0s to NAs. Then for each protein, count the number of samples with non-NA
observation. We can then apply minimum non-missing thresholds
considering the entire sample set, or within
each cohort separately. The latter is worth considering since we plan
to do the analysis in each cohort separately.  

```{r count-nonmissing}
### Convert 0s to NA
prot.study[, names(prot.study) := lapply(.SD, function(x) { x[x==0] <- NA_integer_; x })]


### For each protein, counts the number of samples with non-missing data in each cohort and in total
counts.nonmissing <- data.frame(sapply(unique(ids$source), function(asource) {
    dat <- prot.study[, ids[source == asource, protsample], with = FALSE]
    return(apply(dat, 1, function(x) { sum(!is.na(x)) }))
}))
setDT(counts.nonmissing)
counts.nonmissing[, protein := prot.study$V1]
counts.nonmissing[, wholeset := sum(OHSU, Rush, Emory), by = "protein"]
setcolorder(counts.nonmissing, "protein")


### Determine number of proteins with non-missing data for at least the specified number
### of samples, in each sample set
thresholds <- c(1, 10, 20, 30, 50, 100, 200, 300, n.study)
sets <- setdiff(colnames(counts.nonmissing), "protein")

nproteins.meeting.thresholds <- data.frame(sapply(sets, function(aset) {
    sapply(thresholds, function(x) { sum(counts.nonmissing[, get(aset)] >= x) })
}))


## Add column giving the threshold
nproteins.meeting.thresholds <- data.table(sample.threshold = thresholds, nproteins.meeting.thresholds)


## Add column giving number of proteins meeting threshold in every cohort
Countfunc <- function(athresh) { sum(apply(counts.nonmissing[, cohorts, with = FALSE], 1, function(x) {all(x > athresh)})) }
nproteins.meeting.thresholds[, each.cohort := sapply(thresholds, Countfunc)]

## Get number of proteins with complete data, just for our records
n.proteins.complete.data <- nproteins.meeting.thresholds[sample.threshold == n.study, wholeset]
```
As a reminder, the number of available proteins is `r pN(all.missing["FALSE"])`, 
and the sample sizes for each cohort are: 
`r Mkable(rbind(sort(table(ids$source, useNA = "ifany"), decreasing=TRUE)))`
				   
				   
There are `r pN(n.proteins.complete.data)`
proteins with non-missing data for all `r n.study` samples, but that's a
very stringent threshold. We'll relax the non-missing threshold to
keep more proteins. 

Here are the number of proteins that meet a range of minimum sample thresholds. 


```{r table-protein-keep}
### Draw table of proteins to keep at range of sample thresholds
Mkable(nproteins.meeting.thresholds, caption="number of proteins that would be kept by applying the specified minimum sample threshold to a specific cohort or to the entire sample set")
```

The number of proteins we keep differs in each cohort because the
cohorts have different sample sizes and because the missingness varies by batch and batches are correlated with
cohort. 


```{r define sample threshold}
### Define minimum sample threshold
sampthresh <- 50
```


We're going to keep the 
`r pN(nproteins.meeting.thresholds[sample.threshold == sampthresh, wholeset])` proteins with non-missing observations
for at least `r sampthresh` samples. We'll do PCA to look for sample
outliers. 

First, prepare normalized data by taking log2 of the normalized counts. 
```{r get-log2norm}
### Make vector of proteins to keep and filter data to these proteins
proteins.passed.minn <- counts.nonmissing[wholeset >= sampthresh, protein]
prot.study.minn <- prot.study[V1 %in% proteins.passed.minn, ]

### Extract norm counts to a matrix
### divide each column by column sum to get 
norm0 <- as.matrix(prot.study.minn[, -c("V1")])
norm <- apply(norm0, 2, function(x) { x/sum(x, na.rm=TRUE) })
rownames(norm) <- prot.study.minn$V1

### Take log2 of normalized values
log2norm <- log2(norm)

```

#### Check for duplicate samples
In Kate's QC using the genotype data, she identified two pairs of
duplicate samples. They may or may not be duplicates in this
proteomics data. To check on these pairs, I first calculated the pairwise
Pearson correlation for each pair of samples. Then for each member of
the pair, I calculated the ECDF of the observed correlations with the
other samples. Then I checked the percentile of the observed
correlation between the possible duplicates. 

```{r check-dups}
## Calculate pairwise correlations
cor_matrix <- cor(log2norm, use = "pairwise.complete.obs", method = "pearson")

## Get the sample names and sample indices for pair 1
pair1.names<- ids[ID %in% ids.dups.from.kate[1,], protsample]
pair1.x <- which(rownames(cor_matrix) %in% pair1.names)

## Calculate empirical distribution function usign observed correlations with each sample of interest
## Excluded the correlations for the sample itself and the potential duplicate
## Then check whether the correlation between the potential duplicates falls in the empirical distribution
Ecdf1a <- ecdf(cor_matrix[pair1.x[1], -c(pair1.x)])
Ecdf1b <- ecdf(cor_matrix[pair1.x[2], -c(pair1.x)])
cor_matrix[pair1.x[1], pair1.x[2]]
Ecdf1a(cor_matrix[pair1.x[1], pair1.x[2]])
Ecdf1b(cor_matrix[pair1.x[1], pair1.x[2]])


## Get the sample names and sample indices for pair 2
pair2.names<- ids[ID %in% ids.dups.from.kate[2,], protsample]
pair2.x <- which(rownames(cor_matrix) %in% pair2.names)

## Calculate empirical distribution function usign observed correlations with each sample of interest
## Excluded the correlations for the sample itself and the potential duplicate
## Then check whether the correlation between the potential duplicates falls in the empirical distribution
Ecdf2a <- ecdf(cor_matrix[pair2.x[1], -c(pair2.x)])
Ecdf2b <- ecdf(cor_matrix[pair2.x[2], -c(pair2.x)])
cor_matrix[pair2.x[1], pair2.x[2]]
Ecdf2a(cor_matrix[pair2.x[1], pair2.x[2]])
Ecdf2b(cor_matrix[pair2.x[1], pair2.x[2]])


```
The correlation is pretty high for all the samples. The potential
duplicate pairs did not especially high correlations, so I'm just
going to leave them. 


#### Check on batch effects using PCA
Perform PCA on the log2 values to check for clustering by batch. We
know the missingness varies by batch, but for the PCA we'll use only
the `r pN(n.proteins.complete.data)` proteins with complete data for all samples. 


```{r pca, fig.height=5}
### Extract complete cases for PCA
pca_data <- log2norm[complete.cases(log2norm), ]

### Calculate PCs
pcaObj <-  prcomp(t(pca_data), scale = TRUE)

### Extract PC1 and PC2
## Add sample IDs and batch 
pcpoints <- data.frame(pcaObj$x)[,1:2]
pcpoints$protsample <- rownames(pcpoints)
setDT(pcpoints)
pcpoints <- merge(pcpoints, ids[, .(protsample, Batch)], by = "protsample")

### Add additional labels
## Add column indicating batch groupings (groupings selected by looking at PCA results)
pcpoints[, batchgroup := "batch1-12"]
pcpoints[Batch %in% 16:21, batchgroup := "batch16-21 (6 batches)"]
pcpoints[Batch %in% c(13:15, 22:23), batchgroup := "batch13-15,22-23 (5 batches)"]

## Add column indicating whether sample was excluded from tampor results
pcpoints[, tamporfail := FALSE]
pcpoints[protsample %in% setdiff(pcpoints$protsample, names(tampor)), tamporfail := TRUE]

### Visualize PCs
ggplot(pcpoints, aes(x=PC1, y=PC2, color=Batch, shape = batchgroup)) +
    geom_point() + 
    geom_point(data=subset(pcpoints, tamporfail == TRUE),
               aes(x=PC1, y=PC2), shape = "*", size=7, color="black") +
    ggtitle(paste(nrow(pcpoints), " samples")) +
    theme_minimal() 

```


The samples form three clusters that each contain from five to twelve
batches. 

The Levey group excluded two samples in their post-tampor
dataset. To define outliers, they used WGCNA and picked the samples
with bicor distance sample network connectivity more than three
standard deviations from the mean. 
Their two outliers are both in batch 13 and marked with black asterisks in
the plot.  


										   
```{r finish}

## Add protein column 
log2norm.forprint <- data.table(protein=rownames(log2norm), log2norm)

write.table(ids, file = out.ids.filtered, row.names = FALSE, quote = FALSE, sep = "\t")
fwrite(log2norm.forprint, file = out.log2norm, row.names = FALSE, quote = FALSE, sep = "\t")

```


### Conclusions / To Do 
 * What abundance thresholds do we want to use to define which
   proteins to keep?
 
