---
title: "detect outliers in proteomics data"
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
library(ggplot2)
library(ggrepel)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=6, fig.height=6)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }

RunPCA <- function(pca_data, sdthresh=4) {
    ## pca_data is a data.table with the data (no protein column, just the data columns)
    pcaObj = prcomp(t(pca_data), scale = T)

    #eigs <- pcaObj$sdev^2
    #eigs/ sum(eigs)

    pcpoints = data.frame(pcaObj$x)[,1:2]
    pcpoints$sample=rownames(pcpoints)
    setDT(pcpoints)
    #pcpoints=merge(pcpoints, phenos, all.x = TRUE, by="sample")


    ### Define the limits outside of which a sample will be considered an outlier
    buffer.pc1 <- sdthresh*sd(pcpoints$PC1)
    buffer.pc2 <- sdthresh*sd(pcpoints$PC2)

    keeplims.pc1 <- c(mean(pcpoints$PC1) - buffer.pc1, mean(pcpoints$PC1) + buffer.pc1)
    keeplims.pc2 <- c(mean(pcpoints$PC2) - buffer.pc2, mean(pcpoints$PC2) + buffer.pc2)
    

    ### Initialize columns with NAs
    pcpoints[, c("pc1_outlier", "pc2_outlier", "outlier") := NA_character_]

    ### Fill in sample names for outliers 
    pcpoints[(PC1 > keeplims.pc1[2]) | (PC1 < keeplims.pc1[1]), pc1_outlier := sample]
    pcpoints[(PC2 > keeplims.pc2[2]) | (PC2 < keeplims.pc2[1]), pc2_outlier := sample]
    pcpoints[(!is.na(pc1_outlier) | !is.na(pc2_outlier)), outlier := sample]
    
    return(pcpoints)
}
```


```{r constants}
regressvars <- c("Batch")
sdthreshold <- 5
```

```{r define-input}
outdir <- "2_Pipeline/01-Explore_data/"
file.ids <- paste0(outdir, "ids_semi-filtered.txt")
file.log2norm <- paste0(outdir, "prot_before_sample_outlier_removal_log2norm.txt")
file.resid <- paste0(outdir, "prot_before_sample_outlier_removal_resid_regress", paste0(regressvars), ".txt")

```

```{r define-output}
out.ids.filtered <- paste0(outdir, "ids_pcafiltered.txt")
out.log2norm.pcafiltered <- paste0(outdir, "prot.pcafiltered.log2norm.txt")
```


```{r read-data}
resid <- fread(file.resid)
ids <- fread(file.ids)
log2norm <- fread(file.log2norm)
```

```{r prepare-data-1}



```
## Goal 
Identify outlier samples in the proteomics data using principal
component analysis. 


## Data and Methods
Run the PCA on residuals after regressing out these variables from the
log2 of the normalized abundance profiles: 
`r paste(regressvars, sep=",")`  

Use only the proteins with complete data. 
Define outliers as samples that are more than `r sdthreshold` standard
deviations from the mean for PC1 or PC2. 
We'll iteratively remove outliers and run PCA until no outliers
remain. 

## Results

### Run PCA
```{r do-pca}
### set up for iterative PCA
pcpoints <- vector(mode="list")
pcplots1 <- vector(mode="list")
people_to_drop <- NULL
outlier_count <- -1
pcaround <- 0

### Extract complete data, because that's what PCA requires
forpca <- resid[, !c("protein")]
forpca <- forpca[complete.cases(forpca), ]


### do iterative PCA
while(length(people_to_drop) > outlier_count) {
    ### calculate PCA and identify outlier samples and append to list
    pcaround <- pcaround + 1
    outlier_count <- length(people_to_drop) ## number of outliers found before this round
    pcpoints[[pcaround]] <- RunPCA(forpca[, setdiff(colnames(forpca), people_to_drop), with = FALSE ], sdthreshold)
    
    ## samples that are not outliers have value NA in the PCA object's "outlier" column
    people_to_drop <- c(people_to_drop, as.character(na.omit(pcpoints[[pcaround]]$outlier)))

    ### plot first two PCs
    forplot <- pcpoints[[pcaround]] ## for convenience
    
    pcplots1[[pcaround]] <- ggplot(data = forplot, aes(x = PC1, y = PC2)) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +  # Vertical line at x = 0
        geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +   # Horizontal line at y = 0
        geom_point(color="darkolivegreen") +
        geom_text_repel(aes(label = outlier), color = "red", na.rm = TRUE) +
        theme_bw() +
        ggtitle(paste0("iterative PCA for outlier removal - round ", pcaround)) +
        theme(legend.position="bottom")
}
```
The PCA was done using the `r pN(nrow(forpca))` proteins with complete
data for the 
`r ncol(forpca)` samples. 


Here are the plots from each round of PCA. The outliers, if any, are
labeled in red. 

```{r plot-pca}
for(i in 1:length(pcpoints)) {
    print(pcplots1[[i]])
}
rm(i)

```

After `r pcaround` rounds of PCA we found `r length(people_to_drop)`
outliers.   



### Check for clustering by covariates
Here are PCA plots colored by some potential influential covariates. 

```{r color-plot, fig.width=7}
### Color by covariates of interest
## Use last set of PCs
forplot <- merge(ids, pcpoints[[length(pcpoints)]], by.x = "protsample", by.y = "sample")


### Add additional labels
## Add column indicating batch groupings (groupings selected by looking at PCA results before regressing batch)
forplot[, batchgroup := "batch1-12"]
forplot[Batch %in% 16:21, batchgroup := "batch16-21 (6 batches)"]
forplot[Batch %in% c(13:15, 22:23), batchgroup := "batch13-15,22-23 (5 batches)"]

ggplot(data = forplot, aes(x = PC1, y = PC2, color = factor(msex))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +  # Vertical line at x = 0
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +   # Horizontal line at y = 0
    geom_point() +
    theme_bw() +
    ggtitle("Check clustering by sex") +
    theme(legend.position="bottom")

ggplot(data = forplot, aes(x = PC1, y = PC2, color = factor(source), shape = factor(source))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +  # Vertical line at x = 0
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +   # Horizontal line at y = 0
    geom_point() +
    theme_bw() +
    ggtitle("Check clustering by source") +
    theme(legend.position="bottom")

ggplot(data = forplot, aes(x = PC1, y = PC2, color = factor(batchgroup), shape = factor(batchgroup))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +  # Vertical line at x = 0
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +   # Horizontal line at y = 0
    geom_point() +
    theme_bw() +
    ggtitle("Check clustering by batch group") +
    theme(legend.position="bottom")

ggplot(data = forplot, aes(x = PC1, y = PC2, color = factor(Batch))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +  # Vertical line at x = 0
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +   # Horizontal line at y = 0
    geom_point() +
    theme_bw() +
    ggtitle("Check clustering by batch") +
    theme(legend.position="bottom")

```


```{r filter-log2norm}
### Make new log2norm object with outlier samples removed
log2norm.filtered <- log2norm[, setdiff(colnames(log2norm), people_to_drop), with = FALSE]

```

```{r finish}
### Make new ids object with outlier samples removed
ids.forprint <- ids[protsample %in% names(log2norm.filtered), ]

write.table(ids.forprint, file = out.ids.filtered, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(log2norm.filtered, file = out.log2norm.pcafiltered, row.names = FALSE, quote = FALSE, sep = "\t")
```



## Conclusion / To Do

 * `r length(people_to_drop)` samples were identified as outliers,
   leaving `r ncol(log2norm.filtered)-1` samples for analysis.
 * Even though we regressed out batch, there does still seem to be some
   effect of batch, which also corresponds to cohort. I guess the SVAs
   will account for some of this? 
   
