---
title: "Extract phenotypes for NPS DE - OHSU"
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
#library(ggplot2)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r constants}
sourceset <- "OHSU"

## Define nps variable names -- all 9 domains
npsvars.bin.9 <- c(agitation = "agit", anxiety = "anx", apathy = "apa", delusion = "del",
                 depression = "depd", disinhibition = "disn", hallucination = "hall",
                 irritability = "irr", sleep = "nite")

npsvars.sev.9 <- paste0(npsvars.bin.9, "sev")
names(npsvars.sev.9) <- names(npsvars.bin.9)

## Define nps variable names -- combine hall/del and drop disn
npsvars.bin.7 <- c(agitation = "agit", anxiety = "anx", apathy = "apa",
                 depression = "depd", irritability = "irr", psychosis = "psych",
                  sleep = "nite")

npsvars.sev.7 <- paste0(npsvars.bin.7, "sev")
names(npsvars.sev.7) <- names(npsvars.bin.7)
```


```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }

## tabulate distribution of binary variables
MakeNPSbintable <- function(dattab) {
  ## Tabulate
    disttab <- data.table(t(dattab[, sapply(.SD, function(x) { table(x, useNA = "always") }),
                                   .SDcol = npsvars.bin.9]))
  disttab[, nps:= npsvars.bin.9]
  setnames(disttab, c("0", "1", "V3"), c("no", "yes", "NA"))

  ## Add readable NPS names 
  disttab <- merge(disttab, data.table(NPS = names(npsvars.bin.9), nps = npsvars.bin.9), by = "nps")

  ## Add percent cases column
  ## Reorder rows and columns 
    disttab[, percent.cases := sapply((yes / (no + yes)*100), function(x) {
        paste0(sprintf("%.1f",x), "%") })]
  setcolorder(disttab, c("nps", "NPS", "percent.cases"))
  setorder(disttab, yes)
  return(disttab)
}

## tabulate distribution of severity variables
MakeNPSsevtable <- function(dattab) {
  ## Tabulate
    disttab <- data.table(t(dattab[, sapply(.SD, function(x) { table(x, useNA = "always") }),
                                   .SDcol = npsvars.sev.9]))
  disttab[, nps:= npsvars.sev.9]
  setnames(disttab, c("V5"), c("NA"))

  ## Add readable NPS names 
  disttab <- merge(disttab, data.table(NPS = names(npsvars.sev.9), nps = npsvars.sev.9), by = "nps")


  ## Add case/control columns -- define 0/1 as controls and 2/3 as cases
  ## Add percent cases columns
  disttab[, control.sev := apply(.SD, 1, sum), .SDcol=c("0", "1")]
  disttab[, case.sev := apply(.SD, 1, sum), .SDcol=c("2", "3")]
    disttab[, percent.cases := sapply((case.sev / (case.sev + control.sev)*100), function(x) {
        paste0(sprintf("%.1f",x), "%") })]

  ## Reorder rows and columns 
  setcolorder(disttab, c("nps", "NPS"))
  setorderv(disttab, cols=("0"), order = -1 )
  return(disttab)
}   

```


```{r define-input}
file.ids <- "2_Pipeline/01-Explore_data/ids_pcafiltered.txt"

filedir <- "0_Data/SharepointFiles/"

phesets <- c(nps_cirs = "ALL_NPIQ_CIRS_QUESTIONS_042423",
             demog1    = "N205_WINGO_REGISTRY",
             demog2    = "N206_OHSU_withTissue")

files.phenos <-paste0(filedir, "Phenotypes/OHSU/", phesets, ".txt")
names(files.phenos) <- names(phesets)

```


```{r define-output}
outdir <- "2_Pipeline/01-Explore_data/"
out.cleanedphenos <- paste0(outdir, "phenos_cleaned_", sourceset, ".txt")
out.long <- paste0(outdir, "longitudinal_cleaned_", sourceset, ".txt")
out.patho <- paste0(outdir, "patho_", sourceset, ".txt")
out.patho_filtered_cleaned <- paste0(outdir, "patho_", sourceset, "_filtered_cleaned.txt")


## Command to run this script
thisscript <- "1_Code/01-Explore_data/04-extract_phenos_OHSU.rmd"
#rmarkdown::render(thisscript, output_dir=outdir, knit_root_dir=getwd(), intermediates_dir=getwd())


```


```{r read-data}
### Read in data
ids.all <- fread(file.ids)
phenos <- lapply(files.phenos, fread)
```


```{r prepare-data-1}
### Prepare data
## Extract ids for this source
ids <- ids.all[source=="OHSU", ]


## set ID as numeric
ids[, ID := as.numeric(ID)]

## Update column names
setnames(phenos[["nps_cirs"]], "REGbfoNum", "bfonum")
setnames(phenos[["demog1"]], "REGbfoNum", "bfonum")
setnames(phenos[["demog2"]], "BFO", "bfonum")

setnames(phenos[["demog1"]], "Rpatid", "rpatID")
setnames(phenos[["demog2"]], "SampleID", "rpatID")

## Remove duplicate column
phenos[["nps_cirs"]][, VSTseqno := NULL]

```

## Goal
The goal is to extract the NPS and other variables of interest for the
`r nrow(ids)` `r sourceset` participants in the proteomics dataset. 


## Data
#### Phenotypes
Sharepoint directory: 

Common - Dry Lab > Grant_RO1_NPS_in_AD
Relative paths to files: 
 * N206_OHSU_withTissue.xls
 * 2023 phenotypes OSHU > ALL NPIQ CIRS QUESTIONS 042423.xlsx
 * 2023 phenotypes OSHU > N205 WINGO REGISTRY.xlsx


The NPIQ_CIRS table is longitudinal and the other two are
cross-sectional files with diagnosis and demographic info. 

#### Proteomics samples
The proteomics samples are identified by four-digit numeric IDs which are the
OHSU BFONUMs. 


### Results
## Dimensions of complete phenotype files
```{r check_data}
## Check number of observations and participants in each phenotype set
Mkable(rbind(n.rows = sapply(phenos, nrow), 
      n.unique.participants = sapply(phenos, function(x) { length(unique(x$bfonum))})))
```

## Dimensions after filtering to participants in proteomics datasets
Now I filtered to rows for the `r length(unique(ids$ID))` IDs in
the proteomics ID file. 

```{r extract_data}
### Extract data for participants of interest
## Here we'll keep all the available columns
phenos.ana0 <- sapply(phenos, function(x) { x[bfonum %in% ids$ID, ] }, simplify = FALSE)

## Check row and participant counts again
Mkable(rbind(n.rows = sapply(phenos.ana0, nrow), 
            n.unique.participants = sapply(phenos.ana0, function(x) { length(unique(x$bfonum))})))


### Check how many participants are covered by the NPS datasets
ids.in.nps <- unique(phenos.ana0[["nps_cirs"]]$bfonum)
ids.missing.nps <- setdiff(ids$ID, ids.in.nps)

ids.absent.demog1 <- setdiff(ids.missing.nps, phenos[["demog1"]][, bfonum])
ids.nonpiq.demog1 <- phenos[["demog1"]][bfonum %in% ids.missing.nps & npiqcount == 0, bfonum]


```
Of the `r nrow(ids)` participants with proteomics, I can find 
`r length(ids.in.nps)` in the NPS dataset. Of the 
`r length(ids.missing.nps)` missing, I see that 
`r length(ids.absent.demog1)` are missing from
dataset `demog1` and `r length(ids.nonpiq.demog1)` have value `0` for
variable `npiqcount`.
Some do have non-zero for variable `npcount`, but looks like we only
got NPIQ. Here are the IDs (bfonums) for participants absent from `nps_cirs` data: 

```{r missing-nps}
ids.missing.nps
```

The rest of the exploration will focus on the `r length(ids.in.nps)`
participants present in the data.

```{r , include = FALSE}
### Extract data for participants present in NPS dataset
phenos.ana1 <- sapply(phenos.ana0, function(x) { x[bfonum %in% ids.in.nps, ] }, simplify = FALSE)

## Check row and participant counts one more time 
Mkable(rbind(n.rows = sapply(phenos.ana1, nrow), 
            n.unique.participants = sapply(phenos.ana1, function(x) { length(unique(x$bfonum))})))

### Count visits per participant
n.visits <- table(phenos.ana1[["nps_cirs"]]$bfonum)
summ.n.visits <- summary(as.vector(n.visits))
```

### Number of visits per participant

Here is the summary of the number of visits per participant. 

```{r table-nvisits}
Mkable(rbind((summ.n.visits)))
```


### Sanity checks

```{r check-sex}
sanity <- merge(ids, phenos.ana1[["demog1"]][, .(bfonum, Rsex, Rethnic, REGhispanic,
                                                 Age_at_death, Rdxcode)],
                by.x = "ID", by.y = "bfonum")

checks <- vector()
## Check that sex matches
sextab <- table(sanity[, .(msex, Rsex)])
checks["sex"] <- all(c(sextab[0,1], sextab[1,0]) == 0)

## Check that age at death matches
checks["age_death"] <- sum(sanity[, age_death - Age_at_death]) == 0

## Check race and ethnicity
checks["race"] <- all(sanity$Rethnic == "CAUCASIAN")
checks["ethnicity"] <- all(sanity$REGhispanic %in% c("", "NO"))

msg <- ifelse(all(checks), "They all look fine.",
              "Need to check follow up on this -- there is some discrepancy.")

par(mfrow=c(1,2))
PlotCategorical("Rethnic", sanity)
PlotCategorical("REGhispanic", sanity)

```

I checked the race and ethnicity and compared sex and age at death
from the demographic table to the values in the proteomics table. 
`r msg`

`r rm(msg)`

### Check clinical diagnosis
Here is the breakdown of clinical diagnosis. 


```{r }
demog.merged <- Reduce( function(x, y) {
    merge(x, y, by = c("bfonum", "rpatID")) }, phenos.ana1[c("demog1", "demog2")])

dxkey <- data.table(matrix(c(
"INCIPDEM", "Incipient dementia", 
"MIXED", "Mixed dementia, specify causes", 
"POAD", "Poss AD - other complications",
"PRAD", "Probable Alzheimer's disease",
"QUESTDEM", "Questionable dementia"
), byrow = TRUE, ncol = 2))
names(dxkey) <- c("var", "defini")
setDT(dxkey)
setkey(dxkey, "var")

demog.merged[, diagnosis := dxkey[clinical_dx, "defini"]]
Mkable(addmargins(table(demog.merged[, diagnosis], useNA = "ifany")),
       caption = "Clinical diagnosis from demographic table.")

```

Is there a way to rank these if we want/need to? 


### Get last visit NPS
The NPS data take the usual NPI format, with case/control and
severity variables for each NPS symptom. 


```{r get-last-nps}
varmatch <- sapply(toupper(npsvars.bin.9), grep, names(phenos[["nps_cirs"]]), value = TRUE)

nps.long <- phenos.ana1[["nps_cirs"]][, c("bfonum", "rpatID", "VSTage", "VSTseqno",
                                          as.vector(varmatch)), with = FALSE]
setnames(nps.long, varmatch[1, ], npsvars.bin.9)
setnames(nps.long, varmatch[2, ], npsvars.sev.9)


### Fill in 0 for NAs in severity variables
for(avar in npsvars.bin.9) {
    var.bin <- avar
    var.sev <- paste0(avar, "sev")
    nps.long[is.na(get(var.sev)) & get(var.bin) == 0, (var.sev) := 0]
    nps.long[, (var.sev) := factor(get(var.sev), levels=c(0:3))]
}
rm(avar)

### Substitute -2222 missing code with NA
nps.long[, (names(nps.long)) := lapply(.SD, function(x) {x[x == -2222] <- NA; x}), ]

### Extract last visit per participant
setorder(nps.long, "VSTage")
nps.last <- nps.long[, tail(.SD, 1), by = "bfonum"]
```


```{r check-time-to-death}
### Make table to look at time from death to last NPS visit
timetab <- merge(ids[, .(ID, age_death)], nps.last[, .(bfonum, VSTage)],
                 by.x = "ID", by.y = "bfonum")
timeinterval <- timetab[, age_death - VSTage] 
hist(timeinterval, breaks = 40, col = "azure2",
     main = "time between last NPS and death", xlab = "Years")

```

The time between last NPS visit and death is longer than we saw in the
other cohorts. Still, of the `r nrow(timetab)` participants, `r sum(timeinterval < 2)` 
(`r MakeFrac(mean(timeinterval < 2))`%) have last NPS observation within two
years of death. Is this something we want to filter or adjust for? 

Here is the distribution of case/control status and severity at the
last visit for each NPS.


```{r tabulate-NPS}
## Make tables
dist.nps.bin <- MakeNPSbintable(nps.last)
dist.nps.sev <- MakeNPSsevtable(nps.last)

### Print tables
Mkable(dist.nps.bin, caption = "Case/control distribution at last visit. ")
Mkable(dist.nps.sev, caption = "Severity distribution at last visit. ")
```


```{r find.missing.all.nps, include = FALSE}
### Get IDs for people missing all NPS
index.missing.all.nps <- apply(nps.last[, npsvars.bin.9, with = FALSE], 1, function(x) {
    all(is.na(x)) })
ids.missing.all.nps <- nps.last[index.missing.all.nps, bfonum]

### Check if NPS is available for any other visits 
nps.long[bfonum %in% ids.missing.all.nps, ]
```

There are a few participants with missing values at the last visit for
all NPS variables. These participants have other visits in the
dataset, but they have complete NAs for al of those visits too so I
don't see a way to include them. They'll be dropped from analysis. The ids (bfonums) are: 

```{r print.missing.all.nps}
ids.missing.all.nps

```

```{r update-outcome-tables-exclude-samps-all-missing}
dist.nps.bin.updated <- MakeNPSbintable(nps.last[!bfonum %in% ids.missing.all.nps, ])
dist.nps.sev.updated <- MakeNPSsevtable(nps.last[!bfonum %in% ids.missing.all.nps, ])

Mkable(dist.nps.bin.updated,
       caption = paste0("Case/control distribution at last visit ",
                        "(after removing participants with all missing data). "))
Mkable(dist.nps.sev.updated,
       caption = paste0("Severity distribution at last visit ",
                        "(after removing participants with all missing data). "))
```

```{r add-collapsed-vars}
nps.last[, psych := apply(.SD, 1, any), .SDcol=c("hall", "del")]


```


```{r prep-final-dataset}
### Make vector of GUIDs that have non-missing NPS and meet all other criteria
ids.forana <- setdiff(ids.in.nps, ids.missing.all.nps)

### Assemble the variables of interest for those participants from the phenos file
forprint0 <- merge(demog.merged[bfonum %in% ids.forana, .(bfonum, rpatID, clinical_dx)],
                   nps.last[, c("bfonum", npsvars.bin.9, npsvars.sev.9), with = FALSE], by = "bfonum")

### Merge in variables from proteomics file
ids.sub <- ids[ID %in% ids.forana, ]
forprint1 <- merge(ids.sub[, !c("KX.", "channel", "batch", "WorkingID")],
                   forprint0, by.x="ID", by.y = "bfonum")

### Extract pathologic diagnosis
pathdiag <- demog.merged[bfonum %in% ids.forana, .(bfonum, path_dx)]
pathdiag[, dx_cleaned := gsub("1\\)", "1.0\\)", path_dx)]
pathdiag[, dx_cleaned := gsub("2\\)", "2.0\\)", dx_cleaned)]
pathdiag[, dx_cleaned := gsub("3\\)", "3.0\\)", dx_cleaned)]

pathdiag[, dx1 := tstrsplit(dx_cleaned, split = " 2.0)", keep = 1)]
pathdiag[, dx2 := tstrsplit(dx_cleaned, split = " 2.0)", keep = 2)]
pathdiag[, dx2 := tstrsplit(dx2, split = " 3.0)", keep = 1)]
pathdiag[, dx3 := tstrsplit(dx_cleaned, split = " 3.0)", keep = 2)]

```


```{r prep-long-data}
## Subset to samples with some non-missing data
nps.long.forprint0 <- nps.long[bfonum %in% ids.forana]

## add protsample column
nps.long.forprint1 <- merge(ids.sub[, .(ID, protsample)], nps.long.forprint0,
                            by.x = "ID", by.y = "bfonum")

```

```{r update-nps-variables}
## Add collapsed psychosis variable -- last visit
forprint1[(hall== 1 | del == 1), psych := 1]
forprint1[(hall == 0 & del == 0), psych := 0]

## Add collapsed psychosis variable -- longitudinal 
nps.long.forprint1[(hall== 1 | del == 1), psych := 1]
nps.long.forprint1[(hall == 0 & del == 0), psych := 0]

```


### Covariate distributions
Here are the distributions for some of the covariates of interest for
the `r length(ids.forana)` participants to be included in the analysis. 


```{r hist-covars, fig.height=6}
par(mfrow=c(2,2))


##PlotCategorical("GENDER_DESC", phenos.ana[["demographics"]]);
##title("from demographics table", line=0.5)
PlotCategorical("msex", ids.sub, xsrt=0)
PlotCategorical("clinical_dx", forprint1, xsrt = 45); title("at last visit", line = 0.5)
PlotContinuous("age_death", ids.sub)
PlotContinuous("pmi", ids.sub)

```

These people generally died at older ages than the other cohorts. 
There is one person that died before age 40. 


### Gather dementia severity and pathology severity variables for
### comparing to other cohorts

```{r }
## Not going to print these, but have a look at CIRS longitudinal variables
cirs <- copy(phenos$nps_cirs)
setnames(cirs, names(cirs), sapply(names(cirs), function(x) { gsub(" ", "_", x) }))

## Gather variables to include in patho table
formerge <- list(phenos$demog2[, .(bfonum, path_dx)],
                 phenos$demog1[, .(bfonum, REGageAtOnset, REGpathDx)])

patho_merged <- Reduce(function(x, y) { merge(x, y, by = "bfonum", all = TRUE) }, formerge)

## Use both pathdx columns to get complete data
## I already checked that when both columns have values, the values are the same
## But I see that each column has different missing entries
patho_merged[, path_dx_merged := path_dx]
patho_merged[is.na(path_dx), path_dx_merged := REGpathDx]
patho_merged[, c("path_dx", "REGpathDx") := NULL]
```


### Conclusions / To Do 

 * Of the `r nrow(ids)` participants in the proteomics dataset from this cohort,
 we'll keep `r length(ids.forana)` for analysis.
 * I have the data dictionary for clinical and neuropathology
   diagnosis, but not other variables. Would be nice to have the data
   dictionaries, but it's probably okay since the NPS variables are
   familiar and the other variables I could guess.
 * To Do: Do we want to filter or adjust for time between last NPS
   visit and death? 
 * To Do: Do we want to filter on age at death? 
 * To Do: Are there any other covariates that we want to keep? 
 
 
 

```{r finish}
## Write cleaned phenos to file 
write.table(forprint1, out.cleanedphenos,
            row.names = FALSE, quote = FALSE, sep = "\t")


## Write longitudinal observations to file
write.table(nps.long.forprint1, out.long,
            row.names = FALSE, quote = FALSE, sep = "\t")


## write patho / dementia variables to file
write.table(patho_merged, file = out.patho,
            row.names = FALSE, quote = FALSE, sep = "\t")

## write filtered and cleaned patho dx to file
write.table(pathdiag, file = out.patho_filtered_cleaned,
            row.names = FALSE, quote = FALSE, sep = "\t")

```


