---
title: "Extract phenotypes for NPS DE - Rush"
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r  global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(knitr)
library(data.table)
#library(ggplot2)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```


```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }

```

```{r constants}
sourceset <- "Rush"

```


```{r define-input}
### Files that exist
file.ids <- "2_Pipeline/01-Explore_data/ids_pcafiltered.txt"

filedir <- "0_Data/SharepointFiles/"

phesets <- c("rad201_v1", "rad201_v2", "rad20x", "radintr", "Wingo_Data")
phesets.nps <- setdiff(phesets, "Wingo_Data")
files.phenos <-paste0(filedir, "Phenotypes/Rush/", phesets, ".txt")
```


```{r define-output}
### Files to be created
outdir <- "2_Pipeline/01-Explore_data/"
Out.vars <- function(varset) { paste0(outdir, "rush_vars_", varset, ".txt")  }
out.patho <- paste0(outdir, "patho_", tolower(sourceset), ".txt")
```


```{r read-data}
### Read in data
ids.all <- fread(file.ids)
phenos <- lapply(files.phenos, fread)
names(phenos)  <- phesets
```

```{r prepare-data-1}

### Prepare data
## Extract ids for this source
ids <- ids.all[source=="Rush", ]

ids[, ID := as.integer(ID)]
```

## Goal
The goal is to extract the NPS and other variables of interest for the
`r nrow(ids)` `r sourceset` participants in the proteomics dataset. 

## Data
#### Phenotypes
Sharepoint directory: 

Common - Dry Lab > Grant_RO1_NPS_in_AD > 1_Phenotypes_Frozen tissue >
Rush > Phenotypes_02222023


Single timepoint: 

 * rad201_v1
 * rad201_v2
 * Wingo_Data
 
 Wingo_Data has pathology data, and the other two have NPS. 
 
Longitudinal: 

 * rad20x
 * radintr
 
 These both have NPS. 
 
From notes from Greg Klein: rad201 and rad20x were collected via paper
forms, and radintr is newer data after transitioning to a computer
system. 


I found a report that Jiaqi made in March 2023. This has useful links
to files and maps variable names to NPS symptoms. 

#### Proteomics samples
The proteomics samples are identified by projid (numeric identifiers six to
eight digits in length).  


## Results
### Dimensions of complete phenotype files

```{r }
Mkable(rbind(n.rows = sapply(phenos, nrow), 
      n.unique.GUIDs = sapply(phenos, function(x) { length(unique(x$projid))})))

```

### Dimensions after filtering to participants in proteomics datasets
Now I filtered to rows for the `r length(unique(ids$GUID))` GUIDs in
the proteomics ID file. 

```{r extract_data}
### Extract data for participants of interest
## Here we'll keep all the available columns
phenos.ana <- sapply(phenos, function(x) { x[projid %in% ids$ID, ] }, simplify = FALSE)

## Check row and participant counts again
Mkable(rbind(n.rows = sapply(phenos.ana, nrow), 
            n.unique.GUIDs = sapply(phenos.ana, function(x) { length(unique(x$projid))})))


## How many participants of interest are represented in at least one of the NPS tables?
has.nps <- unique(c(sapply(setdiff(phesets, "Wingo_Data"), function(x) { phenos[[x]]$projid } )))

#length(ids$ID %in% has.nps)

```
Of the `r nrow(ids)` participants with proteomics data, `r length(ids$ID %in% has.nps)` are represented in at
least one of the NPS tables. 


### Check cognitive diagnosis


```{r }
Mkable(rbind(addmargins(table(phenos.ana[["Wingo_Data"]]$cogdx, useNA = "ifany"))))
if(all(phenos.ana[["Wingo_Data"]]$cogdx >=2)) {
    msg <- "All of the participants had final cognitive diagnosis of MCI, AD, or other dementia."
} else{
    msg <- "Check the final cognitive diagnosis. Not all participants had final cognitive diagnosis of MCI, AD, or dementia."
}
```

`r msg`

`r rm(msg)`


### Check sex
```{r check-sex}
sex.from.phenos <- phenos.ana[["Wingo_Data"]][, .(projid, msex)]
setnames(sex.from.phenos, "msex", "msex.fromphenos")

checksex <- merge(ids[, .(ID, msex)], sex.from.phenos, by.x="ID", by.y="projid")

if(all(checksex[, msex == msex.fromphenos])) {
  msg <- "The sex matches between the phenotype data and the proteomics data for all participants."
} else {
  msg <- "Check the sex. There is at least one mismatch between the sex in the phenotype data and the sex in the proteomics data."
}

```

`r msg`

`r rm(msg)`



### Check race
race -- 'With which group do you most closely identify yourself?' -->
value '1' indicates 'White'

spanish-- 'Are you of Spanish/Hispanic/Latino origin?' --> 
value '2' indicates 'No'

```{r check-race}
par(mfrow=c(1,2))
PlotCategorical("race", phenos.ana[["Wingo_Data"]])
PlotCategorical("spanish", phenos.ana[["Wingo_Data"]])
```


### Extracting NPS variables

#### rad201v1 and rad201v2 datasets

```{r define-201-vars}
vars201 <- data.frame(matrix(
    c(
        "depression",    "H1SADDAY",
        "anxiety",       "H5ANXOUS",
        "apathy",        "H14SITST",
        "irritability",  "LOSETEMP",
        "delusion",      "SAIDSTOL",
        "hallucination", "MISPROBJ",
        "sleep1",        "H12MIDDL",
        "sleep2",        "H13EARLY"), 
     byrow = TRUE, ncol=2)
)
names(vars201) <- c("nps", "var")
vars201$var <- tolower(vars201$var)
setDT(vars201)

```


```{r }
## Extract NPS from dataset 1 (201v1)0
phenos.extracted1  <- phenos.ana[[1]][, c("projid", "age_int", vars201$var), with = FALSE]
phenos.extracted1[, set := "201v1"]


phenos.extracted2 <- phenos.ana[[2]][, c("projid", "age_int", vars201$var), with = FALSE]
phenos.extracted2[, set := "201v2"]

extracted201 <- rbind(phenos.extracted1, phenos.extracted2)
setnames(extracted201, vars201$var, vars201$nps)
setorder(extracted201, projid)

#length(unique(extracted201$projid))

```

The 201v1 and 201v2 datasets have almost all the same variables (v1
has one extra, h6sympts). But they use different coding (no/yes/NA is
0/1/9 in v1 and 2/1/8or9 in v2). 


A total of
`r  length(unique(extracted201$projid))` unique participants are
represented in these tables. There are 
`r length(intersect(phenos.ana[[1]]$projid, phenos.ana[[2]]$projid))`
participants present in both tables. 

In Jiaqi's report she picked one item for each NPS to tabulate case
counts, but I see in the notes that that was a quick strategy and the
plan was to revise the NPS definitions later. Looking at the codebook,
it does seem like there might be one or a few items per NPS that we
can use to make a case/control designation. 

Here is the mapping of variable to NPS from
Jiaqi's code: 

```{r }
kable(vars201)

```

The missing rates for these variables are pretty low. 


Here is a look at the difference between age at death (from the
proteomics file) and the age at the NPS observation. 
```{r compare-age, include = FALSE}
### Compare age at death from proteomics id file to age at death from the phenos file
age.death <- merge(phenos[[5]][, .(projid, age_death, pmi)], ids[, .(ID, age_death)],
                   by.x = "projid", by.y = "ID", suffixes = c(".id", ".phe"))
age.message <- ifelse(all((age.death[, age_death.id - age_death.phe]) == 0), 
     "The age at death matches between the proteomics ID file and the phenotype file for all samples.", 
     "There are discrepancies between age at death in the proteomics ID file and the phenotype file.")


### Check age at NPS observation versus age at death
ids[, ID := as.numeric(ID)]
merged <- merge(extracted201, ids[, .(ID, age_death, msex)], by.x = "projid", by.y = "ID")

hist(merged[, age_death - age_int], breaks = 40, col="skyblue2", main = "number of years between NPS observation and death", xlab = "years")
rm(merged)


```

Note: `r age.message`

#### rad20x dataset
The variable labels in this dataset are different from the 201v1 and
201v2 labels, but comparing the codebooks I see the questions are the
same or at least overlapping so I could try to line up the variables
with the 201 datasets. 

The variable labels in this dataset use keywords, so I can more easily count
the number of variables per NPS. Here are the keywords that Jiaqi
used, and the number of variables per NPS. 

```{r }
keywords <- c(depression = "depress",
              anxiety = "upset",
              apathy = "apathy",
              irritability = "temper",
              delusion = "trust",
              halluciantion = "delusi",
              sleep5 = "sleep5",
              sleep8 = "sleep8")

x20vars <- sapply(keywords, function(akeyword) { grep(akeyword, names(phenos[[3]]), value = TRUE)  }, simplify = FALSE)

varcounts <- data.table(cbind(nps = names(keywords), keyword = keywords, varcount = sapply(x20vars, length)))
setorder(varcounts, nps)

Mkable(varcounts, caption = "number of variables per NPS based on keywords")
print("\n\n")
```


```{r extract-20x}
### Extract the variables of interest for participants of interest
extracted20x <- phenos.ana[[3]][, c("projid", "age_int", unlist(x20vars)), with = FALSE]


### Check number of observations per participant
obs.per.participant.20x <- as.vector(table(extracted20x$projid))

Mkable(rbind(summary(obs.per.participant.20x)), caption = "Summary of number of visits per participant")
```

This dataset includes `r length(unique(extracted20x$projid))` participants. 
There are up to `r max(obs.per.participant.20x)` visits per participant. 

The data have good completeness (data not shown). 
 
```{r }
### Calculate non-missing rate per variable
## non-missing values take values 1-5 (yes/no or indicating duration or frequency range)
nonmissing <- data.table(cbind(variable = setdiff(names(extracted20x), c("projid", "age_int")),
                               nonmissing.rate = round(apply(extracted20x[, -c("projid", "age_int")], 2, function(x) {
                                   sum(x %in% c(1:5) )})/nrow(extracted20x), 2)))

#kable(nonmissing)

```

```{r }
selectvars20x <- c(depression = "depress",
              anxiety = "upset",
              apathy = "apathy1",
              irritability = "temper1",
              delusion = "trust1",
              halluciantion = "delusi1",
              sleep = "sleep5")        
```

```{r check-visit-ages-rad20x, include=FALSE}
### Check age at NPS observation versus age at death
merged <- merge(phenos.ana[["rad20x"]][, .(projid, age_int)], ids[, .(ID, age_death, msex)], by.x = "projid", by.y = "ID")

hist(merged[, age_death - age_int], breaks = 40, col="skyblue2", main = "number of years between NPS observation and death", xlab = "years")
rm(merged)
```


### radintr dataset
This dataset uses yet a different variable label scheme, but again
seems to use pretty much the same questions so the variables could be
lined up with the other datasets. 

```{r }
selectvarsint <- c(depression = "semdn",
              anxiety = "upset",
              apathy = "sitst",
              irritability = "temper",
              delusion = "hsstl",
              halluciantion = "msprc",
              sleep2 = "wakup")

```

```{r }
extractedint <- phenos[[4]][, c("projid", "age_int", selectvarsint), with = FALSE]


### Check number of observations per participant
obs.per.participant.int <- as.vector(table(extractedint$projid))

Mkable(rbind(summary(obs.per.participant.int)), caption = "Summary of number of visits per participant")
```

This dataset includes `r length(unique(extractedint$projid))` participants. 
There are up to `r max(obs.per.participant.int)` visits per participant. 


```{r check-visit-ages-radint, include=FALSE}
### Check age at NPS observation versus age at death
merged <- merge(extractedint, ids[, .(ID, age_death, msex)], by.x = "projid", by.y = "ID")

hist(merged[, age_death - age_int], breaks = 40, col="skyblue2", main = "number of years between NPS observation and death", xlab = "years")
rm(merged)
```


### Number of visits per participant
Here is the summary of the number of visits per participant after combining
all of the NPS datasets. 
```{r check-number-visits-across-datasets}
### Combine the projid and age_int columns from each nps set so I can check the last available NPS for each person
visits.all <- rbindlist(sapply(phesets.nps, function(x) { phenos.ana[[x]][, .(projid, age_int)]   },
                            simplify = FALSE), idcol = "pheset")


Mkable(rbind(summary(as.integer(table(unique(visits.all[, .(projid, age_int)])$projid)))))
```



### Select last available visit 

```{r check-last-dataset}
## Count the number of individuals in different subsets of datasets
presence_matrix <- data.frame(sapply(phenos[1:4], function(x) { ids$ID %in% x$projid }))
presence_matrix$n.sets <- apply(presence_matrix, 1, sum)
setDT(presence_matrix)
presence_matrix$projid <- ids$ID


Pairwise_comparison <- function(x, y) {
  list(
    shared = intersect(x, y),
    unique_to_x = setdiff(x, y),
    unique_to_y = setdiff(y, x)
  )
}

combinations <- combn(setdiff(phesets, "Wingo_Data"), 2, simplify = FALSE)

pairwise_results <- lapply(combinations, function(pair) {
  list1 <- phenos[[pair[1]]]$projid
  list2 <- phenos[[pair[2]]]$projid
  comparison <- Pairwise_comparison(list1, list2)
  names(comparison) <- c("Shared", paste0("Unique to ", pair[1]), paste0("Unique to ", pair[2]))
  comparison
})

names(pairwise_results) <- sapply(combinations, paste, collapse = " vs ")
```



```{r check-time-death-last-NPS}
### Sort by age at visit and extract last visit per participant
setorder(visits.all, age_int)
lastvisit <- visits.all[, tail(.SD, 1), by = "projid"]
lastvisit <- merge(lastvisit, ids[, .(ID, age_death)], by.x = "projid", by.y = "ID")


hist(lastvisit[, age_death - age_int], breaks=20, col = "azure2",
     main = "Time between last NPS and death", xlab = "Years")

summary(lastvisit[, age_death - age_int])

```




All the participants have an NPS visit within 
`r ceiling(max(lastvisit[, age_death - age_int]))` years of
death. There are 
`r sum(lastvisit[, age_death - age_int] < 0)` participants with
negative time, but that is probably due to sometimes only having years
instead of precise dates.  

Here is the breakdown of which dataset has the last available visit
for our participants. This is before checking for NAs, but to check
for NAs we need to confirm which variables we'll use from each
dataset. 


```{r table-last-pheset}
Mkable(rbind(addmargins(table(lastvisit$pheset))))

```


```{r extract-last-visits}
## For each participant, extract the last available visit
lastobs.list <- sapply(phesets.nps, function(aset) {
    ## Get projids for participants whose last visit was in this set
    projids <- lastvisit[pheset == aset, projid]

    ## Extract all visits for those participants
    extracted <- phenos.ana[[aset]][projid %in% projids, ]

    ## Return last visit for those participants
    setorder(extracted, age_int) ## sort by age at visit
    return(extracted[, tail(.SD, 1), by = "projid"]) ## return last row for each participant
}, simplify = FALSE)


```

```{r count-missings, eval=FALSE}
## To pick variables, we want to know if there are any variables that have more missingness so we
## can exclude them from consideration. Since the different datasets have different variables names
## and I haven't aligned them all yet, I'm just focusing on the participants with their last visit in the rad20x
## dataset, which has the most logical variable names and for which I have mapped the variables to data types
count.missings.lastvisit.20x <- data.frame(n = cbind(apply(lastobs.list[["rad20x"]], 2, function(x) { sum(x %in% 1:6) })))
count.missings.lastvisit.20x$variable <- rownames(count.missings.lastvisit.20x)



varstocheckcorr <- scan("vars_tocheck_corr.txt", what = "") ## read tmp file
tocheckcorr <- lastobs.list[["rad20x"]][, tolower(varstocheckcorr), with = FALSE]

tocheckcorr[tocheckcorr> 2] <- NA_integer_


```


### Check covariates


```{r fig.height=6}
par(mfrow=c(2,2))
PlotCategorical("msex", ids)
PlotCategorical("cogdx", phenos.ana[["Wingo_Data"]])
PlotContinuous("age_death", ids)
PlotContinuous("pmi", ids)


```


```{r write-vars}
for(x in names(phenos)) {
    write(names(phenos[[x]]), file = Out.vars(x), ncolumns = 1)
}
rm(x)
```


### Gather dementia severity and pathology severity variables for
### comparing to other cohorts

```{r }

patho  <- phenos.ana$Wingo_Data[, .(projid, cogdx, ad_reagan, niareagansc,
                                hip_scl_3reg_yn, hspath_typ, caa_4gp, cvda_4gp2, ci_num2_gct, ci_num2_mct)]

                                         
```


```{r finish}
write.table(patho, file = out.patho, row.names = FALSE, quote = FALSE, sep = "\t")

```


## To do
 * Line up variables across datasets
 * Define NPS case/control criteria using available variables and
   longitudinal data
