---
title: "Extract Phenotypes for NPS DE - Emory"
date: "`r Sys.Date()`"
author: "Selina Vattathil"
output: word_document
---
```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```
```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
library(ggplot2)
library(lubridate)
options(stringsAsFactors = FALSE)
```

```{r constants}
sourceset <- "emory"

## Define nps variable names -- all 9 domains
npsvars.bin.9 <- c(agitation = "agit", anxiety = "anx", apathy = "apa", delusion = "del",
                 depression = "depd", disinhibition = "disn", hallucination = "hall",
                 irritability = "irr", sleep = "nite")

npsvars.sev.9 <- paste0(npsvars.bin.9, "sev")
names(npsvars.sev.9) <- names(npsvars.bin.9)

## Define nps variable names -- combine hall/del and drop disn
npsvars.bin.7 <- c(agitation = "agit", anxiety = "anx", apathy = "apa",
                 depression = "depd", irritability = "irr", psychosis = "psych",
                  sleep = "nite")

npsvars.sev.7 <- paste0(npsvars.bin.7, "sev")
names(npsvars.sev.7) <- names(npsvars.bin.7)
```


```{r functions}
source("~/comsv/Svattathil_Library/svattathil_functions.r")
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }

```


```{r define-input}
### Files that exist
filedir <- "0_Data/SharepointFiles/"

## The phenotypes are divided into multiple files
phesets1 <- c("demographics", "diagnosis", "historyofmed", "mocammse", "npi")
phesets2 <- c("UDS1", "UDS2", "UDS3")
phesets3 <- c("dob")
files.phenos <- c(paste0(filedir, "Phenotypes/Emory/bhc_findinggenetics_",
                         phesets1, "_bds866.csv"),
                  paste0(filedir, "Phenotypes/Emory/", phesets2, "-DR2023421WingoNPS.txt"),
                  paste0(filedir, "Phenotypes/Emory/", "guids_n28_dob.txt"))

names(files.phenos) <- c(phesets1, phesets2, phesets3)

## this file is from Jiaqi's lookup in OBIEE
file.supprace <- paste0(filedir, "Phenotypes/Emory/guids_w_race_n35.txt")

file.ids <- "2_Pipeline/01-Explore_data/ids_pcafiltered.txt"

file.idmap <- paste0(filedir, "Phenotypes/Emory/case_number_GUID_map.txt")
```


```{r define-output}
outdir <- "2_Pipeline/01-Explore_data/"
out.cleanedphenos <- paste0(outdir, "phenos_cleaned_", sourceset, ".txt")
out.long <- paste0(outdir, "longitudinal_cleaned_", sourceset, ".txt")
out.patho <- paste0(outdir, "patho_", sourceset, ".txt")

## Command to run this script
thisscript <- "1_Code/01-Explore_data/06-extract_phenos_emory.rmd"
#rmarkdown::render(thisscript, output_dir=outdir, knit_root_dir=getwd(), intermediates_dir=getwd())


```


```{r read_data}
ids.all<- fread(file.ids)
phenos <- sapply(files.phenos, fread, simplify = FALSE)
idmap <- fread(file.idmap, header=TRUE)
supprace <- fread(file.supprace)
```


```{r prepare_data}
## Update column names
setnames(idmap, "Case_Number", "casenumber")
setnames(phenos[["mocammse"]], "GLOBAL_SUBJECT_ID", "GUID")

## Restrict ids to this source
ids0 <- copy(ids.all[source=="Emory", ])


### Add GUID to proteomics ids
ids <- merge(ids0, idmap[, .(casenumber, GUID)],  by.x="ID", by.y="casenumber")


### Harmonize UDS visit year columns
## Thanks to Jiaqi for mapping these out
dates1 <- as.Date(phenos[["UDS1"]][, em_visit_date], format = "%m/%d/%y")
phenos[["UDS1"]][, VISITYR := as.numeric(format(dates1, "%Y"))]
setnames(phenos[["UDS2"]], "visit year", "VISITYR")
setnames(phenos[["UDS3"]], "visityr", "VISITYR")

setnames(phenos[["UDS1"]], "em_visit_seq_no", "VISITNUM")
setnames(phenos[["UDS2"]], "em_visit_seq_no", "VISITNUM")
setnames(phenos[["UDS3"]], "visitnum", "VISITNUM")


### Set DOB as date in each of the two tables with DOB
## have to make dates.fixed because otherwise years are assumed to be 20XX instead of 19XX
phenos[["dob"]][, dates.fixed := sub("^(\\d{1,2})/(\\d{1,2})/(\\d{2})$", "\\1/\\2/19\\3", DOB_default)]
phenos[["demographics"]][, dates.fixed := sub("^(\\d{1,2})/(\\d{1,2})/(\\d{2})$", "\\1/\\2/19\\3", DOB)]

## Convert to date format
phenos[["dob"]][, dob1 := as.Date(dates.fixed,  format = "%m/%d/%Y")]
phenos[["demographics"]][, dob1 := as.Date(dates.fixed,  format = "%m/%d/%Y")]


### Add race observations from Jiaqi's OBIEE lookup. This is for participants that had missing race
### data in the demographics table.
setnames(supprace, "Race Default", "RACE_DESC")
setnames(supprace, "Global Subject ID", "GUID")

for(aguid in supprace[, GUID]) {
    newrace <- supprace[GUID == aguid, RACE_DESC]
    if(aguid %in% phenos[["demographics"]][, GUID]) {
        ## replacing existing value
        phenos[["demographics"]][GUID == aguid, RACE_DESC := newrace]
    } else {
        ## add new row to demographics -- will have data for only GUID and race
        phenos[["demographics"]] <- rbindlist(list(phenos[["demographics"]], supprace[GUID == aguid, ]),
                                              fill=TRUE)
    }
}
rm(aguid)

```

## Goal
The goal is to extract the NPS and other variables of interest for the
`r nrow(ids)` participants in the proteomics dataset. 


## Data
#### Phenotypes
Sharepoint directory: 

Common - Dry Lab > Grant_RO1_NPS_in_AD > 3_Emory NPS data

I'm using seven phenotype files. Five different phenotype files with
suffix _bds866.csv from directory 'data from Emory': 

Cross-sectional: 

 * demographics

 
Longitudinal: 

 * diagnosis
 * historyofmed
 * mocammse
 * npi

I am referencing Jiaqi's reports from October 2023, which referenced
these files. 

I'm also using the file UDS1-3_DR2023421WingoNPS.xlsx from directory
'data from GADRC'. This excel file has a sheet with a map of case
number, GUID, and CRIN ID, and three NPS datasets:

Longitudinal:

 * UDS1
 * UDS2
 * UDS3 
 
Finally, Jiaqi queried Emory OBIEE 
for DOB for 28 participants that were
missing from the demographics file above, and race for those 28
participants plus 7 participants that did not report race. 
 * dob
 * race


#### Proteomics samples
The proteomics samples are identified by case IDs of format E12-34 or
OS-123456. The phenotype files are identified by GUID or patient
ID. To link them, I used the table in the GADRC UDS1-3 excel sheet.


## Results
### Merge UDS1 followup variables
The variables are mostly the same across the datasets, except UDS1 separates the followup
visits from the initial visits, i.e. there is variable `normcog` and
separate variable `fu_normcog`. For all the variables of interest,
I combined the initial and followup observations by filling in missings
in the initial with the observation from the `fu` variable. That will
make the variable names consistent across datasets.


```{r merge-UDS1-fu-variables}
## Make vector of variables that need to be merged with their fu_counterparts
## variable fu_newinf doesn't have a matching newinf variable
vars.tomerge <- substring(grep("^fu", names(phenos[["UDS1"]]), value = TRUE), 4)
vars.tomerge <- setdiff(vars.tomerge, "newinf")

## Do merge/replacement 
for(avar in vars.tomerge) {
    col.ini <- avar               
    col.fu <- paste0("fu_", avar)
    
    phenos[["UDS1"]][is.na(get(col.ini)), (col.ini) := get(col.fu)]
}

```

### Dimensions of complete phenotype files
```{r check_data}
## Check number of observations and participants in each phenotype set
## demographics is cross-sectional, but there are a few duplicated participants
## the other phenotypes are longitudinal
Mkable(rbind(n.rows = sapply(phenos, nrow), 
      n.unique.GUIDs = sapply(phenos, function(x) { length(unique(x$GUID))})))
```

### Dimensions after filtering to participants in proteomics datasets
Now I filtered to rows for the `r length(unique(ids$GUID))` GUIDs in
the proteomics ID file. 

```{r extract_data, include = TRUE}
### Extract data for participants of interest
## Here we'll keep all the available columns
phenos.ana <- sapply(phenos, function(x) { x[GUID %in% ids$GUID, ] }, simplify = FALSE)

## Check row and participant counts again
Mkable(rbind(n.rows = sapply(phenos.ana, nrow), 
            n.unique.GUIDs = sapply(phenos.ana, function(x) { length(unique(x$GUID))})))


### Check how many participants are covered by the NPS datasets
guids.in.nps <- unique(unlist(sapply(c("npi", phesets2), function(x) { phenos.ana[[x]]$GUID })))

guids.in.UDS <- unique(unlist(sapply(c(phesets2), function(x) { phenos.ana[[x]]$GUID })))

### Check how many are only in the NPI dataset
#setdiff(guids.in.nps, unique(unlist(sapply(c(phesets2), function(x) { phenos.ana[[x]]$GUID }))))


```

Of the `r nrow(ids)` participants with proteomics, I can find `r length(guids.in.UDS)` in the UDS
datasets. 

Regarding the one missing, in the list of 93 people noted as having NPS in
the UDS1-3 -DR2023421WingoNPS.xlsx file, I see that GUID 40588 is
highlighted in red as 'not a UDS participant', so there must have been
some issue there. I am guessing that NPS is not available for this
person. 
  

### Sanity checks
#### Demographics table
The phenotype data includes a demographic table, but only 
`r length(unique(phenos.ana[["demographics"]]$GUID))` of the `r nrow(ids)`
participants of interest are included. While I can get sex from the
proteomics file, I need the demographics file to get date of birth,
and I need date of birth if we want age at last NPS observation. The
visits in the UDS dataset are identified by visit date, not
participant age. 


#### Sex
Check that msex from
proteomics ID table matches gender description from demographics
phenotype file:

```{r check_sex}
merged <- merge(ids, phenos[["demographics"]], by = "GUID")
sextab <- table(merged[, .(msex, GENDER_DESC)])
nmismatch <- sextab[1,2] + sextab[2,1]
Mkable(table(merged[, .(msex, GENDER_DESC)]))

guids.sexmismatches <- merged[(msex == 1 & GENDER_DESC=="Female") |
                        (msex == 0 & GENDER_DESC=="Male"), GUID]

```

Note: there is `r nmismatch` mismatch between the proteomics `msex`
value and the demographics table `GENDER_DESC` value among the `r sum(sextab)`
participants. 


#### Race
Here is the reported race from the demographics table. 
`r Mkable(rbind(addmargins(table(phenos.ana[["demographics"]]$RACE_DESC, useNA="ifany"))))`


```{r }
## Count the number of individuals in different subsets of datasets
phesets.nps <- c("npi", phesets2)
presence_matrix <- data.frame(sapply(phenos[phesets.nps], function(x) { ids$GUID %in% x$GUID }))
presence_matrix$n.sets <- apply(presence_matrix, 1, sum)
setDT(presence_matrix)
presence_matrix$GUID <- ids$GUID

```

```{r get-last}
### Combine the GUID and VISITYR and VISITNUM columns from each UDS set
### to count number of visits and identify the last available NPS for each person
visits.all <- rbindlist(sapply(phesets2, function(x) { phenos.ana[[x]][, .(GUID, VISITYR, VISITNUM)]   },
                            simplify = FALSE), idcol = "pheset")

### Count visits per participant
n.visits <- table(visits.all$GUID)
summ.n.visits <- summary(as.vector(n.visits))

### Sort by year and visit number and extract last visit per participant
setorder(visits.all, GUID, VISITYR, VISITNUM)
lastvisit <- visits.all[, tail(.SD, 1), by = "GUID"]
lastvisit <- merge(lastvisit, ids[, .(GUID, age_death)], by = "GUID")


```


### Number of visits per participant
Here is the summary of the number of visits per participant. 

```{r table-nvisits}
Mkable(rbind((summ.n.visits)))
```

### Determine NPS dataset with last observation for each participant
Here is the breakdown of which dataset has the last available visit
for our participants. This is before checking for NAs. 

```{r table-last-visit-pheset}
Mkable(rbind(addmargins(table(lastvisit$pheset))))
```


```{r extract-last-visits}
## For each participant, extract the last available visit
lastobs1 <- sapply(phesets2, function(aset) {
    ## Get projids for participants whose last visit was in this set
    guids <- lastvisit[pheset == aset, GUID]

    ## Extract all visits for those participants
    extracted <- phenos.ana[[aset]][GUID %in% guids, ]

    ## Return last visit for those participants
    setorder(extracted, VISITNUM) ## sort by visit number
    return(extracted[, tail(.SD, 1), by = "GUID"]) ## return last row for each participant
}, simplify = FALSE)


```


### Get last visit cognitive status
```{r get-last-visit-cog-status}
## Extract columns
fordx <- rbindlist(list(UDS1 = lastobs1[["UDS1"]][, .(GUID, VISITYR, VISITNUM, normcog, demented, probad)],
                        UDS2 = lastobs1[["UDS2"]][, .(GUID, VISITYR, VISITNUM, normcog, demented, probad)],
                        UDS3 = lastobs1[["UDS3"]][, .(GUID, VISITYR, VISITNUM, normcog, demented, alzdis)]),
                   use.names = TRUE, fill = TRUE, idcol="pheset")

fordx[, dx := NA_character_]
fordx[normcog == 1, dx := "normal"]
fordx[normcog == 0 & (demented == 1 | probad == 1 | alzdis == 1), dx := "AD"]
fordx[normcog == 0 & demented == 1 & (probad == 0 | alzdis == 0), dx := "other_dementia"]

```

```{r check-moca-mmse, include = FALSE}
## Extract rows for our participants with non-missing result
mocammse.nonmissing <- phenos[["mocammse"]][(GUID %in% ids$GUID) &
                                            !is.na(SERVICE_DT) &
                                            (!is.na(MOCA_RAW_SCORE) |
                                             !is.na(MMSE_RAW_SCORE)), ]

## count participants
n.mocammse <- length(unique(mocammse.nonmissing$GUID))

## check if participants with normal cognition have moca/mmse scores
fordx[dx == "normal", GUID] %in% mocammse.nonmissing$GUID

```

Jiaqi used a combination of variables `normcog`, `demented`, `probad`, and
`alzdis` to define the dementia status for each participant. 
Using the same criteria, here is the cognitive status distribution at
the last visit: 

`r Mkable(rbind(addmargins(table(fordx$dx, useNA="ifany"))))` 

Two participants had normal cognition at their last visit, but from
other notes we know these had dementia at their final diagnosis, so
they will be kept in the analysis. 

We do have a dataset with MOCA/MMSE scores, but we have non-missing
data for only `r n.mocammse` participants. 



```{r eval=FALSE}
sapply(phenos[phesets2], names)
```


### Time between last NPS and death
```{r time-last-visit-death}
### Aggregates dates of birth from the two phenos tables
dobs0 <- rbind(phenos.ana[["demographics"]][, .(GUID, dob1)],
              phenos.ana[["dob"]][, .(GUID, dob1)])

### Merge in date of last visit
### Add version with month and day, assume January 1
lastage <- merge(dobs0, fordx[, .(GUID, VISITYR)], by = "GUID")
lastage[, visityr := as.Date(paste0(VISITYR, "-01-01"))]

### Calculate age at last visit
lastage[, age.int := lubridate::time_length(interval(dob1, visityr), "years")]

### Calculate time between last NPS visit and death
timetab <- merge(lastage, ids[, .(GUID, age_death)], by = "GUID")
timeinterval <- timetab[, age_death - age.int]

hist(timeinterval, breaks=20, col="Azure2",
     main = "Time between last NPS and death", xlab = "Years")

```

Of the `r nrow(timetab)` participants, `r sum(timeinterval < 2)` 
(`r MakeFrac(mean(timeinterval < 2))`%) have last NPS observation within two
years of death. Is this something we want to filter or adjust for? 


### Get last visit NPS status 
I'm assuming we are interested in nine individual NPS symptoms. For
each NPS, there is a binary variable and a severity variable. Here are
the distributions of each. 


```{r get-last-visit-nps}
### Extract last visit NPS -- both case/control and severity measures
fornps <- rbindlist(list(UDS1 = lastobs1[["UDS1"]][, c("GUID", "VISITYR", "VISITNUM",
                                                       npsvars.bin.9, npsvars.sev.9), with = FALSE],
                         UDS2 = lastobs1[["UDS2"]][, c("GUID", "VISITYR", "VISITNUM",
                                                       npsvars.bin.9, npsvars.sev.9), with = FALSE],
                         UDS3 = lastobs1[["UDS3"]][, c("GUID", "VISITYR", "VISITNUM",
                                                       npsvars.bin.9, npsvars.sev.9), with = FALSE]),
                   use.names = TRUE, fill = TRUE, idcol="pheset")


### Tabulate case/control status
MakeNPSbintab <- function(dattab) {
    ## Tabulate
    dist.nps.bin <- data.table(t(dattab[, sapply(.SD, function(x) { table(x, useNA = "always") }),
                                        .SDcol = npsvars.bin.9]))
    dist.nps.bin[, nps:= npsvars.bin.9]
    setnames(dist.nps.bin, c("0", "1", "V3"), c("no", "yes", "NA"))
  
    ## Add readable NPS names 
    dist.nps.bin <- merge(dist.nps.bin, data.table(NPS = names(npsvars.bin.9), nps = npsvars.bin.9),
                          by = "nps")

    ## Add percent cases column
    ## Reorder rows and columns 
    dist.nps.bin[, percent.cases := sapply((yes / (no + yes)*100), function(x) {
        paste0(sprintf("%.1f",x), "%") })]
    setcolorder(dist.nps.bin, c("nps", "NPS", "percent.cases"))
    setorder(dist.nps.bin, percent.cases)
    return(dist.nps.bin)
}

dist.nps.bin <- MakeNPSbintab(fornps)


### Tabulate severity status
## Save original severity variables with suffix .orig
## We are going to recode them from NA/1/2/3 to 0/1/2/3
for(avar in npsvars.bin.9) {
    var.bin <- avar
    var.sev <- paste0(avar, "sev")
    fornps[, (paste0(var.sev, ".orig")) := get(var.sev)]
    fornps[is.na(get(var.sev)) & get(var.bin) == 0, (var.sev) := 0]

    fornps[, (var.sev) := factor(get(var.sev), levels=c(0:3))]
}
rm(avar)

## Tabulate
dist.nps.sev <- data.table(t(fornps[, sapply(.SD, function(x) { table(x, useNA = "always") }),
                                    .SDcol = npsvars.sev.9]))
dist.nps.sev[, nps:= npsvars.sev.9]
setnames(dist.nps.sev, c("V5"), c("NA"))

## Add readable NPS names 
dist.nps.sev <- merge(dist.nps.sev, data.table(NPS = names(npsvars.sev.9), nps = npsvars.sev.9),
                      by = "nps")

## Reorder rows and columns 
setcolorder(dist.nps.sev, c("nps", "NPS"))
setorderv(dist.nps.sev, cols=("0"), order = -1 )

### Print tables
Mkable(dist.nps.bin, caption = "Case/control distribution at last visit. ")

Mkable(dist.nps.sev, caption = "Severity distribution at last visit. ")

```

There are a few participants with missing values at the last visit for
all NPS variables. These all had only one available visit, and they
had AD. These will be dropped from analysis. GUIDs:


```{r find.missing.all.nps}
index.missing.all.nps <- apply(fornps[, c(npsvars.bin.9), with = FALSE], 1, function(x) { all(is.na(x)) })
ids.missing.all.nps <- fornps[index.missing.all.nps, GUID]
ids.missing.all.nps

```
```{r }
dist.nps.bin.updated <- MakeNPSbintab(fornps[!GUID %in% ids.missing.all.nps, ])
Mkable(dist.nps.bin.updated,
       caption = paste0("Case/control distribution at last visit ",
                        "(after removing participants with all missing data)."))

```


```{r prep-final-dataset}
### Make vector of GUIDs that have non-missing NPS and meet all other criteria
ids.forana <- setdiff(ids$GUID, ids.missing.all.nps)
ids.forana <- intersect(ids.forana, guids.in.nps)

### Assemble the variables of interest for those participants from the phenos file
forprint0 <- fornps[, c("GUID", npsvars.bin.9, npsvars.sev.9), with = FALSE]

### Merge in variables from proteomics file
ids.sub <- ids[GUID %in% ids.forana, ]
forprint1 <- merge(ids.sub[, !c("KX.", "channel", "batch", "WorkingID")], forprint0,
                   by.x="GUID", by.y = "GUID")

```


### Covariate distributions
Here are the distributions for some of the covariates of interest for
the `r length(ids.forana)` participants to be included in the analysis. 

```{r hist-covars, fig.height=6}
par(mfrow=c(2,2))

##PlotCategorical("GENDER_DESC", phenos.ana[["demographics"]]);
##title("from demographics table", line=0.5)
PlotCategorical("msex", ids.sub)
PlotCategorical("dx", fordx[GUID %in% ids.forana,], xsrt=45); title("at last visit", line = 0.5)
PlotContinuous("age_death", ids.sub)
PlotContinuous("pmi", ids.sub)

```


```{r prep-long-data}
datecols <- c("DAY_SERVICE_DATE", "em_visit_date", "VISITYR", "VISITNUM")
datecols <- c("em_visit_date", "VISITYR", "VISITNUM")

forlong <- sapply(phesets2, function(aset) {
    phenos[[aset]][GUID %in% ids.forana,
                   intersect(c("GUID", datecols, npsvars.bin.9, npsvars.sev.9), names(phenos[[aset]])),
                   with = FALSE]
    }, simplify = FALSE)


nps.long.forprint0 <- rbindlist(forlong, idcol="pheset", fill = TRUE)
setorder(nps.long.forprint0, GUID, VISITYR, VISITNUM)
nps.long.forprint1 <- merge(ids.sub[, .(protsample, GUID)], nps.long.forprint0, by = "GUID")

```

```{r update-nps-variables}
## Add collapsed psychosis variable -- last visit
forprint1[(hall== 1 | del == 1), psych := 1]
forprint1[(hall == 0 & del == 0), psych := 0]

## Add collapsed psychosis variable -- longitudinal 
nps.long.forprint1[(hall== 1 | del == 1), psych := 1]
nps.long.forprint1[(hall == 0 & del == 0), psych := 0]

```


### Gather dementia severity and pathology severity variables for
### comparing to other cohorts
```{r }
## Extract last available moca/mmse scores
mocammse_long = phenos.ana$mocammse[, .(GUID, SERVICE_DT, TEST_NM, MOCA_RAW_SCORE, MMSE_RAW_SCORE), ]                       
setorder(mocammse_long, SERVICE_DT)
mocammse_last <- mocammse_long[, tail(.SD, 1), by = GUID]

## Add prefix labels to columns for each dataset
setnames(mocammse_last, setdiff(names(mocammse_last), "GUID"),
         paste0("score_", setdiff(names(mocammse_last), "GUID")))
setnames(fordx, setdiff(names(fordx), "GUID"), paste0("UDS_", setdiff(names(fordx), "GUID")))


## Merge with last available UDS diagnosis
patho <- merge(mocammse_last, fordx, by = "GUID", all = TRUE)

                  
```




### Conclusions / To Do 

 * Of the `r nrow(ids)` participants in the proteomics dataset from this cohort,
 we'll keep `r nrow(ids.sub)` for analysis.
 * To Do: Is there any way to resolve the one sample with sex discrepancy?
 * To Do: Can we verify that NPS for GUID 40588 is not available?
 * To Do: Are there any other covariates that we want to keep? 
 


```{r finish}
## Write cleaned phenos to file
write.table(forprint1, file = out.cleanedphenos, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(nps.long.forprint1, file = out.long, row.names = FALSE, quote = FALSE, sep = "\t")
write.table(patho, file = out.patho, row.names = FALSE, quote = FALSE, sep = "\t")
```
