---
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
library(ggplot2)
library(lavaan)
library(semPlot)
library(psych)
library(polycor)
library(argparse)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r constants}
options(stringsAsFactors = FALSE)
source("1_Code/project_constants.r")
```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }
```


```{r command-line}
### Command line arguments
## Create parser
parser <- ArgumentParser()
parser$add_argument("--max_factors", type="integer", default=4, help="Maximum number of factors to test")

## Read from parser
args <- parser$parse_args()

currcohort <- "rush"

```

---
title: "Factor analysis - `r currcohort`"
---



```{r define-input}
### Files that exist
indir <- "2_Pipeline/02-Prepare_analysis_data/"
infiles <- list(phenos = paste0(indir, "phenos_cleaned_SVs_", currcohort, ".txt")
              )
```


```{r define-output}
### Files to be created
outdir <- "2_Pipeline/02-Prepare_analysis_data/Latent_factors"
MyMkdir(outdir)
outfiles <- list(factor_scores  = file.path(outdir, paste0("factor_scores_", currcohort, ".txt")),
                 rds            = file.path(outdir, paste0("efa_obj_", currcohort, ".rds"))
                 )

## Command to run this script
thisscript <- "1_Code/02-Prepare_analysis_data/02b-estimate_latent_factors_rush.rmd"
#rmarkdown::render(thisscript, output_dir=outdir, knit_root_dir=getwd(), intermediates_dir=getwd())
```


```{r read-data}
phenos <- fread(infiles[["phenos"]])

```


```{r prepare-data-1}
## Convert to matrix for analysis
nps_matrix <- as.matrix(phenos[, npsvars.bin, with = FALSE])
rownames(nps_matrix) <- phenos$protsample


# Ensure binary coding (0/1)
if (!all(nps_matrix %in% c(0, 1, NA))) {
  cat("WARNING: Non-binary values detected. Converting to 0/1...\n")
  nps_matrix <- ifelse(nps_matrix > 0, 1, 0)
}


# Check for complete cases
complete_cases <- complete.cases(nps_matrix)
cat("Complete cases:", sum(complete_cases), "out of", nrow(phenos), "\n")

if (sum(complete_cases) < nrow(nps_matrix)) {
  cat("Proceeding with listwise deletion for missing data.\n")
  nps_matrix <- nps_matrix[complete_cases, ]
}

```


```{r exploratory-factor-analysis}
### Compute tetrachoric correlation matrix
tetra_corr <- psych::tetrachoric(nps_matrix)$rho

```


#### Determine number of latent factors
```{r scree}
### Determine number of factors using scree plot
screeout <- fa.parallel(tetra_corr, n.obs = nrow(nps_matrix), fm = "minres", fa = "fa")

```
The suggested number of latent factors based on the screen plot is `r screeout$nfact`. 

#### EFA model fit - `r screeout$nfact` latent factors
```{r fit-efa-nfact, echo=TRUE}
## Use rotate = "oblimin" if you believe factors may be correlated (likely in psych data).
## Another rotation method is 'varimax'
## Use cut = 0.2 to suppress very low loadings in the printout.

```{r fit-efa, echo=TRUE}
## Use rotate = "oblimin" if you believe factors may be correlated (likely in psych data).
## Another rotation method is 'varimax'
## Use cut = 0.2 to suppress very low loadings in the printout.
efa_results <- list(rep(NA, 4))

for(curr_n in 1:4) {
efa_results[[curr_n]] <- psych::fa(tetra_corr,
                          nfactors = curr_n,
                          n.obs = nrow(nps_matrix),
                          rotate = "oblimin",
                          fm = "minres")
}

print(efa_results[[screeout$nfact]], digits = 2, cut = 0.2)
```


```{r print-visualize-loadings-nfact}
## Factor loadings table
cat("\nComplete factor loadings:\n")
loadings_table <- unclass(efa_results[[screeout$nfact]]$loadings)
loadings_df <- as.data.frame(loadings_table)
loadings_df$Variable <- rownames(loadings_df)
loadings_df$h2 <- efa_results[[screeout$nfact]]$communality
print(round(loadings_df[, setdiff(colnames(loadings_df), "Variable")], 3))

## Visualization of loadings
fa.diagram(efa_results[[screeout$nfact]], main = paste0(screeout$nfact, "-factor structure -", currcohort))

```


#### Interpretation -- 1-factor model 
Basic notes on interpreting latent factor results:

 * The latent factors have names MR* because they were extracted using
   the 'minimal residual' method
 * loadings: indicate which factor a symptom belongs to. 
    + More technically, they describe how much of the variance in the observed
   variable is explained by the factor. They are the regression
   coefficient of the observed variable on the factor. 
    + loadings > 0.4 are strong loadings 
    + Ideally each factor should have 2+ strong loadings. Single-item
     factors indicate overfitting. The factors could reflect unique
     variance or noise, not true latent constructs
    + We shouldn't expect the loadings we're inspecting in the pattern
     matrix to sum to 1. The squared loadings for a variable across
     all factors plus the uniqueness (residual variance) should sum to 1 if you're 
	 working with standardized data. But since our model allows
     factors to correlation (oblimin rotation), that won't necessarily
     hold. 
 * communalities (h2): How much of a symptom’s variance is explained
   by the latent factors.
   + Values > 0.4 are decent.
   + Value of 1 (no unexplained residual variance) is rare and suggests overfitting
 * Uniqueness (u2) = 1 - h2: Residual variance not explained.
 * Item complexity -- reflects how many factors a variable loads onto.
   + Low (≈ 1.0) = Simple structure
   + Higher (≥ 2.0) = variable loads across multiple factors
 * SS Loadings: Sum of squared loadings for each factor. Describes the
 total amount of variance in all variables that the factor accounts
 for.
 * Proportion Var: proportion of total variance explained by each
   factor
 * Cumulative Var: proportion of variance explained by all factors
 * Proportion Explained: How much of the explained variance is due to
   each factor
 * RMSR: closer to 0 = better
 * Test of model vs. null = Objective function; 0 = better
 * Factor score adequacy: how well the factor scores would work in
   future modeling
 

#### EFA model fit - `r 2` latent factors

As a sensitivity test, I fit a 2-factor model. 
```{r fit-efa-2, echo=TRUE}
## Use rotate = "oblimin" if you believe factors may be correlated (likely in psych data).
## Another rotation method is 'varimax'
## Use cut = 0.2 to suppress very low loadings in the printout.
curr_n <- 2
efa_results[[curr_n]] <- psych::fa(tetra_corr,
                              nfactors = curr_n,
                              n.obs = nrow(nps_matrix),
                              rotate = "oblimin",
                              fm = "minres")
print(efa_results[[curr_n]], digits = 2, cut = 0.2)
fa.diagram(efa_results[[curr_n]], main = paste(curr_n, "- factor structure -", currcohort))
```


```{r get-scores}
factors_to_use <- 1

factor_scores_obj <- psych::factor.scores(x = nps_matrix,
                                          f = efa_results[[factors_to_use]],
                                          method = "regression")
factor_scores <- data.table(factor_scores_obj$scores, keep.rownames = "protsample")
```

Here is the distribution of  each factor score. 

```{r visualize-score-distributions, fig.height = 8}
factorcols <- setdiff(colnames(factor_scores), "protsample")
nfactors <- length(factorcols)

par(mfrow=c(nfactors, 1))
invisible(sapply(factorcols, function(afactor) {
    hist(unlist(factor_scores[, afactor, with = FALSE]),
         main = afactor, xlab = "", breaks = 30)
}))


```


```{r finish}
write.table(factor_scores, file = outfiles[["factor_scores"]],
            row.names = FALSE, quote = FALSE, sep = "\t")

saveRDS(efa_result_nfact, file = outfiles[["rds"]])

```
