---
author: "Selina Vattathil"
date: "`r Sys.Date()`"
output: word_document
---

```{r global-options-1, include=FALSE}
rm(list=ls())
library(knitr)
library(data.table)
library(ggplot2)
library(argparse)
library(dplyr)
library(tidyr)
library(gridExtra)
options(stringsAsFactors = FALSE)
source("~/comsv/Svattathil_Library/svattathil_functions.r")
```

```{r global-chunk-settings-1, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=TRUE, message=TRUE, include=TRUE, collapse=TRUE)
```

```{r word-document-settings-1, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(fig.width=7, fig.height=4)
```

```{r html-document-settings-1, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r constants}
npsvars.bin <- c(agitation = "agit", anxiety = "anx", apathy = "apa", delusion = "del",
                 depression = "depd", disinhibition = "disn", hallucination = "hall",
                 irritability = "irr", sleep = "nite")

```

```{r functions, include=FALSE}
## function to use , thousands separator for numbers in kable
Mkable <- function(x, ...) { kable(x, format.args = list(big.mark = ','), ...) }
```


```{r command-line-arguments}
## Create parser
parser <- ArgumentParser()
parser$add_argument("--cohort", type="character", default = "OHSU", help="OHSU or emory")

## Read from parser
args <- parser$parse_args()


```

---
title: Explore longitudinal data -- `r args$cohort`
---

```{r define-input}
infiles <- list(long = paste0("2_Pipeline/01-Explore_data/longitudinal_cleaned_", args$cohort, ".txt"),
              basic = paste0("2_Pipeline/01-Explore_data/phenos_cleaned_", args$cohort, ".txt"))
```


```{r define-output}
outfiles <- list(persistent_cases = paste0("2_Pipeline/01-Explore_data/phenos_persistent_cases_", args$cohort, ".txt"))
```


```{r read-data}
dat.basic <- fread(infiles[["basic"]])
dat.long <- fread(infiles[["long"]])

```


```{r prepare-data-1}

```

### How many visits per participant?
Note: These counts are not considering whether the visit has real
observations or NA. The NAs may vary across NPS. 
```{r }
visits.pp <- dat.long[, list(n.visits = .N), by = "protsample"]
visits.pp.tab <- table(visits.pp$n.visits)

vis_df <- data.frame(
  visits   = as.integer(names(visits.pp.tab)),   # number of visits (x‑axis)
  participants = as.integer(visits.pp.tab),     # count of participants (y‑axis)
  stringsAsFactors = FALSE
)

## Print summary statistics
summary(visits.pp$n.visits)

## Draw bar plot
ggplot(vis_df, aes(x = factor(visits), y = participants)) +
  geom_col(fill = "#4C72B0", width = 0.7) +           # same colour as base R bar
  geom_text(aes(label = participants), vjust = -0.5, size = 3.5) +  # label counts
  labs(
    title    = paste0("Number of visits per participant (", args$cohort, ", N = ", sum(visits.pp.tab), ")"),
    x        = "Visits per participant",
    y        = "Participants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )

```


```{r non-na-visits, eval = FALSE}
### How many visits with data per participant, per NPS? 
for(anps in npsvars.bin) {
    this.nps <- dat.long[!is.na(unlist(dat.long[, anps, with = FALSE])), ] ## visits with non-missing for this nps
    this.nps.tab <- table(this.nps[, list(n.good.visits = .N), by = "protsample"]$n.good.visits)

    print(anps)
    print(kable(rbind(this.nps.tab), row.names = FALSE, caption = anps))
}
rm(anps)

```


### How many visits have NA observation, per NPS?
```{r check-missingness}
kable(dat.long[, lapply(.SD, function(x) { sum(is.na(x)) }), .SDcol = npsvars.bin])
```


### Plot the participant's case rate against their number of visits
```{r plot-case-rate,  fig.width=7, fig.height=10}
 
n.case.visits <- dat.long[, lapply(.SD, function(x) { sum(x==1, na.rm=TRUE) }), .SDcols = npsvars.bin, by = "protsample"]
n.control.visits <- dat.long[, lapply(.SD, function(x) { sum(x==0, na.rm=TRUE) }), .SDcols = npsvars.bin, by = "protsample"]
n.total.visits <- dat.long[, lapply(.SD, function(x) { sum(x %in% c(0,1), na.rm=TRUE) }), .SDcols = npsvars.bin, by = "protsample"]

setorder(n.case.visits, protsample)
setorder(n.total.visits, protsample)


if(all(n.case.visits$protsample == n.total.visits$protsample)) {
    mean.across.visits <- n.case.visits[, -c("protsample")] / n.total.visits[, -c("protsample")]
    
}

### Plot the participant's case rate across visits against their number of visits
## Set up for the plot
max_visits <- max(unlist(n.total.visits[, -c("protsample")]))

## This is a helper table to draw the lines 
curve_grid <- crossing(
  x = 1:max_visits,   # every integer x you care about
  n = 1:max_visits              # each constant n
)
setDT(curve_grid)
curve_grid <- curve_grid[n <= x, ]
curve_grid[, y := n / x]

## Draw a plot for each NPS and save as a list
rate_plots <- sapply(npsvars.bin, function(currnps) {
    toplot <- cbind(total_visits = unlist(n.total.visits[, ..currnps]), mean_status = unlist(mean.across.visits[, ..currnps]))

    ## Draw lines for 0 case visits, 1 case visit, 2 case visits, etc.
    p <- ggplot(toplot, aes(x = total_visits, y = mean_status)) +
        geom_line(data = curve_grid,
            aes(x = x, y = y, group = factor(n)), colour = "grey80", 
            linewidth = 0.5, linetype = "dashed", show.legend = FALSE) +
        geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.5, linetype = "dashed") + 

    ## Add jittered points; adjust width/height as needed
        geom_jitter(width = 0.15, height = 0.005,
                    shape = 21,               # filled circle
                    colour = "NavyBlue",
                    fill   = "DodgerBlue",
                    size   = 2,
                    alpha  = 0.6) +           # semi‑transparent for overlap
        labs(title = paste0(names(npsvars.bin)[npsvars.bin == currnps]),
             x = "Total Visits",
             y = "Rate of case visits") +
    theme_minimal() +
    theme(panel.grid.major.y = element_blank(),
                  panel.grid.minor.y = element_blank()) +
    scale_x_continuous(breaks = 1:max_visits) +
    scale_y_continuous(breaks = seq(0, 1, 0.2))

    return(p)
}, simplify = FALSE)


## Print the plots 
marrangeGrob(grobs = rate_plots, layout_matrix = matrix(c(1:9, NA), ncol = 2, byrow = TRUE), top = "")


```



### How many participants are cases at >=2 visits? 


Definitions:  
Note -- a visit is counted for an NPS only if it had a non-NA observation. 

 * case:       participant had at least two visits and was case at at least 2 visits,
 * control:    participant had at least two visits and was control at all visits,
 * NA:         participant had only one visit, or was case at exactly one visit.


```{r get-persistent-cases}
## Combine total visit counts and case visit counts into one table
merged_dt <- merge(
  n.total.visits,
  n.case.visits,
  by = "protsample",
  suffixes = c(".tot", ".case"),
  all = TRUE   
)

## Apply the three‑rule logic (vectorised)
## Make a list with one element per NPS
status_list <- lapply(npsvars.bin, function(var) {
    tot_col  <- merged_dt[[paste0(var, ".tot")]]   # extract for convenience
    case_col <- merged_dt[[paste0(var, ".case")]]  # extract for convenience
    
    status <- fifelse(
        tot_col >= 2 & case_col >= 2, 1,       # case -- ≥2 total visits AND ≥2 case visits
        fifelse(
            tot_col >= 2 & case_col == 0, 0,   # control -- ≥2 total visits AND 0 case visits
            NA_integer_                        # everything else → NA
        )
    )
    return(status)
})

## Use code names and convert to data table with sample column
names(status_list) <- npsvars.bin
status_cols <- as.data.table(status_list)
nps.persist <- data.table(protsample = merged_dt$protsample, status_cols)


kable(nps.persist[, lapply(.SD, function(x) { sum(x, na.rm = TRUE) }),  .SDcol = npsvars.bin],
      caption = "Case counts")

kable(nps.persist[, lapply(.SD, function(x) { sum(is.na(x)) }),  .SDcol = npsvars.bin],
      caption = paste0("Total participants dropped from analysis (out of ", nrow(nps.persist), ")"))



## Make new phenos table using the persistent case definition
## first copy the existing phenotype table
## and delete the old NPS columns
dat.persist0 <- copy(dat.basic)
dat.persist0[, c(npsvars.bin, paste0(npsvars.bin, "sev")) := NULL]

## Add new nps columns
dat.persist <- dplyr::left_join(dat.persist0, nps.persist, by = "protsample")

## Old definition -- case if case at >=2 visits, control otherwise
#nps.persist <- dat.long[, lapply(.SD, function(x) { ifelse((sum(x, na.rm = TRUE) >= 2), 1, 0) }), .SDcol = npsvars.bin, by = "protsample"]

```


For comparison, here are the case counts from the last visit data we
have been using:

```{r print-current}
kable(dat.basic[, lapply(.SD, function(x) { sum(x) }),  .SDcol = npsvars.bin],
      caption = "Case counts when case = 'case at last visit'")
```


### Compare average to last visit NPS 
```{r compare-average}
nps.avg <- dat.long[, lapply(.SD, function(x) { mean(x, na.rm = TRUE)  }), .SDcol = npsvars.bin, by = "protsample"]

setorder(nps.avg, protsample)
setorder(dat.basic, protsample)


if (!identical(dat.basic$protsample, nps.avg$protsample)) {
  stop("The 'protsample' columns are not identical or not in the same order.")
}

## Initiate vectors to hold results
corr_vals <- numeric(length(npsvars.bin))
p_vals    <- numeric(length(npsvars.bin))

for (i in seq_along(npsvars.bin)) {
    col_name <- npsvars.bin[i]

    x <- dat.basic[[col_name]]
    y <- nps.avg[[col_name]]

    
    test_res <- cor.test(x, y, method = "pearson")
    corr_vals[i] <- test_res$estimate   # correlation coefficient
    p_vals[i]    <- test_res$p.value    # two‑sided p‑value
}


symptom_corr <- data.table(
  symptom     = npsvars.bin,
  correlation = round(corr_vals, 3),
  p_value     = signif(p_vals, 3)
)

# Create an empty matrix filled with NA
corr_mat <- matrix(NA_real_, nrow = length(npsvars.bin), ncol = length(npsvars.bin),
                   dimnames = list(npsvars.bin, npsvars.bin))

# Fill the diagonal with the actual correlation values
diag(corr_mat) <- symptom_corr$correlation

## reshape to long format
corr_long <- as.data.table(reshape2::melt(corr_mat,
                                          varnames = c("symptom_x", "symptom_y"),
                                          value.name = "correlation"))

ggplot(corr_long, aes(x = symptom_x, y = symptom_y, fill = correlation)) +
    geom_tile(color = "white", na.rm = FALSE) +               # draws the squares

    ## Print the numeric value only on the diagonal (where we have data)
    geom_text(aes(label = ifelse(!is.na(correlation),
                                 sprintf("%.2f", correlation), "")),
              size = 4, colour = "black") +

    ## Diverging colour scale centred at zero
    scale_fill_gradient2(low  = "#4C72B0",   # blue for negative
                         mid  = "white",
                         high = "#D55E00",   # orange/red for positive
                         midpoint = 0,
                         limits = c(-1, 1),
                         na.value = "grey95",   # colour for empty cells
                         name = "Correlation") +

                                        # Clean up axes
    labs(title = paste0("Correlation of last-visit and average NPS -- ", args$cohort,
                        "\n(using binary observations)"),
         x = NULL, y = NULL) +

    theme_minimal(base_size = 12) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank(),
        legend.position = "right"
    )
```


```{r finish}
### Write new phenotype file using persistent case definition to file
fwrite(dat.persist, file = outfiles[["persistent_cases"]], sep = "\t")
```

